---
title: Working with APIs and Functions
---


## What is an API!?

To put it simply, An API^[[IBM's definition](https://www.ibm.com/think/topics/api)] 
or **application programming interface** can be thought of in part as another 
tool in the user's tool belt to *programmatically* interact and often extract 
data without having to manually download data. No one needs to lift their
mouse and search through a webpage and click download.  This is a simple,
yet powerful concept that extend to many areas of data work and beyond. 
I encourage users to read up on and develop their API skills beyond 
what is presented in our coursework.


## Working with `httr2`

We will be working with [`httr2`](https://httr2.r-lib.org/articles/httr2.html)
as our package of choice for API work. Often, specific APIs will have their
own packages for working with them, it is often better to use those, but we
cannot cover them all, and not every API has its own package. `httr2` will
always be here for us, and if users can understand `httr2` then picking
up the other more specific packages will become even easier.

Remember our Covid-19 pandemic dataset? Before we would navigate to that 
website and download the datafile, or perhaps rely on the snippet of data
provided in this course. No more - we are far too busy for that kind of work.
From here on out we will be using the API, and in the future when users
encounter data they want to incorporate into their analyses - it is almost
always better to use the API.

Install `httr2` like we would any other package.

```{r}
#| label: installing httr2
#| eval: false

install.packages('httr2')
```

Next we'll attach the `httr2` and `dplyr` packages to our session.

```{r}
#| label: httr2-intro-libs
library(dplyr) |> suppressPackageStartupMessages() # this silences dplyr messages when loading the package
library(httr2)
```


### `httr2` requests and responses

Recall our [pandemic data's landing page](https://data.ct.gov/Health-and-Human-Services/COVID-19-Tests-Cases-and-Deaths-By-Town-ARCHIVE/28fr-iqnx/about_data) -
notice what pops up when you click the 'Export' button in the top right of
the webpage. The popup has 'Download file' defaulting to 'Export format' of
csv. However, right next to it is 'API endpoint' - click that and notice
the 'Data format' switches to 'JSON'. Below that you may notice that
there is a 'Version' toggle for either 'SODA3' or 'SODA2'. For now, select
'SODA2' and copy the url provided, it should look like this:
[https://data.ct.gov/resource/28fr-iqnx.json](https://data.ct.gov/resource/28fr-iqnx.json)
now that we have our url or **endpoint** as they are often called let's hop
into the code. 


```{r}
#| label: httr2-request-response

# save our url into an object so we don't have to type it out everywhere
endpoint <- "https://data.ct.gov/resource/28fr-iqnx.json"

# We build our data request up piece by piece

my_request <- request(endpoint) # base of the request

# Get a our response from the Open Data Portal API

odp_response <-
  my_request |>
  req_perform() # perform the request, and get a response base

odp_response
```


### Extracting data from responses

We have a response - now where is our data?  We need to pull it out of
the body of the response and then convert it into an R dataframe.


```{r}
#| label: httr2-data

my_odp_data <-
  odp_response |>
  resp_body_json() |> # extract the data, it's returned as an R list
  bind_rows() # for this endpoint we can easily convert to a dataframe

my_odp_data
```


Lovely - but how come there are only 1000 rows? That leads us to our next
wrinkle when working with APIs. This one is public, so just about anyone
can hammer on this API at anytime, and some of data can be quite large. In 
order to prevent users from accidentally running a massive query for data they
may not even need (and to prevent misuse from unscrupulous types) APIs will
often only send a small chunk of the data and then offer up ways to retrieve
the next chunk, and then the next chunk, and so on and so forth.  This can
be accomplished in a few different ways, but for this API they use pages
with **limits** and **offsets**. 


::: {.callout-note}
Users working with different APIs should read *that* API's 
documentation to see how that specific API handles pagination/iteration
through responses.
:::




## Extracting *all* data

We're going to have to do a little leg work here to get *all* the
data, and then I am going to show you how you can package this all up
so you can reuse it over and over again.

First, users need to understand what the limit and offset are and
how they can leverage those to get the data they expect.

- **limit**: The total number of *rows* of data brought back in the request.
    - The default for *this* API and this endpoint is 1000. Some datasets might
    be small like 5000 rows. You can just bump this number up to 5000 and pull
    it all in one go.

- **offset**: The starting point row number for the query.
    - The default for *this* API is 0, because it starts at the first row.

These two parameters can be used in conjunction to, for example, grab the 
first 1000 rows of a dataset, and then grab the next 1000 and so on and so
forth until all data is had.

### Modifying the limit and offset

So modifying the limit and offset is all well in good, but how do we
actually go about changing those values? Remember our endpoint url?
Thankfully these are parameters that can be tacked on to the end
of the url.

If for whatever reason we wanted the 10th row, and *only* the 10th row
of our dataset we can modify our url to look like this:

[`https://data.ct.gov/resource/28fr-iqnx.json?$limit=1&$offset=9`](https://data.ct.gov/resource/28fr-iqnx.json?$limit=1&$offset=9)

Notice that the parameters come at the end of the original url after a `?`.
Users then append `$limit=` and `&` and `$offset=` with the desired values.
Feel free to try this in your web browser - most modern browsers will display
the data from the response. 


::: {.callout-note}

Note that items/data are zero-indexed, meaning the first item the 
data is 'row' 0. To grab the 10th item we skip the first 9 items.

:::


Do not take my word for it. Try this out:
click on [`https://data.ct.gov/resource/28fr-iqnx.json?$limit=10&$offset=0`](https://data.ct.gov/resource/28fr-iqnx.json?$limit=10&$offset=0)
and make note of the tenth item.

It should look something like this:

```json
{
    "lastupdatedate":"2022-06-24T00:00:00.000",
    "town_no":"10","town":"Bethlehem",
    "towntotalcases":"620",
    "townconfirmedcases":"529",
    "townprobablecases":"91",
    "towncaserate":"18225",
    "towntotaldeaths":"3",
    "townconfirmeddeaths":"2",
    "townprobabledeaths":"1",
    "peopletested":"2781",
    "ratetested100k":"81746",
    "numberoftests":"11841",
    "numberofpositives":"749",
    "numberofnegatives":"11070",
    "numberofindeterminates":"22"
}
```

Does that match our original single item we had before?

### Using httr2

No one wants to modify urls by hand, and we're still clicking through 
everything, time to use `httr2` to do the heavy lifting.


```{r}
# We're going to grab first 5000 rows, 1000 rows at a time.
# so we will query 1000 rows, 5 times

req <-
  request(endpoint) |>
  # don't forget the $ as part of the parameter name for this api
  # backticks are required for nonstandard r names
  req_url_query(
    `$limit` = 1000,
    `$offset` = 0
  )


resps <-
  req_perform_iterative(
    req,
    next_req = iterate_with_offset(
      param_name = "$offset",
      start = 0,
      offset = 1000
    ),
    max_reqs = 5
  )

# resps contains a list of 5 responses
# we want to take each response and do what we did above to all of them
# we will use an anonymous function otherwise known as lambda functions in 
# other programming languages

data <- 
    resps |> 
    resps_data(
        \(resp) # declare the name of each item in our resps list
            resp |> # for each response in our responses
            resp_body_json() |> # convert the json to a list
            bind_rows() # bind these into a dataframe
    )

data
```

### All the data everywhere, all at once

Wait - weren't we promised all the data!?
Fair point, how do we grab that data if we don't know how many rows there are?
At time of writing there are 102245 rows in this dataset, but let's pretend
we do not know that, and often that is the case.

```{r}

# bump this up to 10k rows per request and 10k offset each times
number_of_rows = 10000

# first request grabs the first 10k
req <-
  request(endpoint) |>
  req_url_query(
    `$limit` = number_of_rows,
    `$offset` = callout-warning0
  )

# drop the max_reqs argument
resps <-
  req_perform_iterative(
    req,
    next_req = iterate_with_offset(
      param_name = "$offset",
      start = 0,
      offset = number_of_rows
    )
  )

data <- 
    resps |> 
    resps_data(
        \(resp) 
            resp |> 
            resp_body_json() |> 
            bind_rows()
    )

nrow(data) == 102245
```

Look at that! We have our 102245 rows. 


::: {.callout-warning}

***Start small when working with a new endpoint***

Users will often want *all* the data, but they will often *not* know how many 
rows there are. Use caution when grabbing everything, there may be more data
than can be handled at once.

:::


