---
title: "Implementing APIs into our Analysis"
execute:
  # Live API calls - see the note in this part's lesson chapter.
  freeze: true
---

The Commissioner's next question is bigger than one town: **how many
tests has Connecticut run in total, as of the most recent report?**
Rather than re-downloading a CSV every time that number is needed,
this script pulls the freshest data straight from the API and
calculates the answer itself - the same `fetch_all_data()` function
from the lesson, now put to work.

## Setup

Same packages as the lesson, plus `lubridate` this time, to parse the
timestamp the API sends back.

```{r}
#| label: setup
library(dplyr)
library(httr2)
library(lubridate)
```

## Step 1 — Hoist the constants

The resource ID is the one piece of this script most likely to
change - next year's report might live under a different four by
four. Constants like this get pulled to the top of the script in
SCREAMING_CASE by convention, so anyone skimming the file can spot
what's configurable at a glance, instead of hunting through function
bodies for a hardcoded string.

```{r}
#| label: constants
RESOURCE <- "28fr-iqnx"
```

## Step 2 — The fetch function

This is the same `fetch_all_data()` function from the lesson, dropped
in as-is. Keeping the fetch logic in its own function - rather than
writing the request/response/pagination dance directly into the
script - means the messy details of pagination live in exactly one
place. If the API changes tomorrow, there's one function to fix, not
every script that happens to need CT COVID data.

```{r}
#| label: fetch-function
fetch_all_data <-
  function(resource, rows = 10000) {
    # Fetch all data from the provided resource id for the CT Open Data Portal
    #
    # resource: string - Required portal "four by four" in the format 'xxxx-xxxx'
    # rows: numeric - Number of rows of data to fetch one response at a time.
    #   defaults to 100000 and becomes the LIMIT and OFFSET for the request.
    # Returns: A tibble

    # handle the resource from the user

    base <- "https://data.ct.gov/resource"

    # create the first request
    req <-
      request(base) |>
      req_url_path_append(resource) |> #append the user provided 4x4 to the base url
      req_url_query(
        `$limit` = rows, # use the default rows=10000 or what ever the user provided
        `$offset` = 0 # start with the first row
      )

    # perform the request and then with each request increment the offset by `rows`
    resps <-
      req_perform_iterative(
        req,
        progress = FALSE,
        next_req = iterate_with_offset(
          param_name = "$offset",
          start = 0,
          offset = rows # use the default rows=10000 or what ever the user provided
        )
      )

    # convert the response objects from the API to a tibble of data
    # for each response from the API grab the json body, convert to an R list
    # and then squash all that into a singular tibble
    data <-
      resps |>
      resps_data(
        \(resp) {
          resp |>
            resp_body_json() |>
            bind_rows()
        }
      )

    return(data)
  }
```

## Step 3 — A `main()` entrypoint

Nothing has actually run yet - so far this script has only defined a
constant and a function. `main()` is where the analysis itself lives:
fetch the data, then compute the one number the Commissioner asked
for. Wrapping a whole analysis in a function called `main()` (a
convention borrowed from other programming languages) buys you two
things: one place to read top-to-bottom to see the entire analysis at
a glance, and a script that does nothing until you actually call it.
That second part matters once a script gets sourced by other scripts -
you don't want an API call firing off just because something else
needed one function out of the file.

```{r}
#| label: main-function
main <- function() {
  # main entrypoint into our analysis

  # data ingestion
  data <- fetch_all_data(resource = RESOURCE)

  # key measures
  total_tests <-
    data |>
    mutate(date = ymd_hms(lastupdatedate)) |>
    filter(date == max(date)) |>
    reframe(
      total_pandemic_tests = sum(
        as.numeric(numberoftests),
        na.rm = TRUE
      )
    ) |>
    pull(total_pandemic_tests)

  cat(
    paste(
      "The total number of pandemic tests in these data were:",
      format(total_tests, big.mark = ',')
    )
  )
}
```

## Step 4 — Running the analysis

Defining `main()` didn't run anything - calling it does.

```{r}
#| label: run-main
main()
```

## Putting it together

Each step above built up one script piece by piece. Written as a
single file, the whole thing looks like this:

```{r}
#| eval: false

# === setup ===
library(dplyr)
library(httr2)
library(lubridate)

# === constants ===
RESOURCE <- "28fr-iqnx"

# === functions ===
fetch_all_data <-
  function(resource, rows = 10000) {
    # Fetch all data from the provided resource id for the CT Open Data Portal
    #
    # resource: string - Required portal "four by four" in the format 'xxxx-xxxx'
    # rows: numeric - Number of rows of data to fetch one response at a time.
    #   defaults to 100000 and becomes the LIMIT and OFFSET for the request.
    # Returns: A tibble

    # handle the resource from the user

    base <- "https://data.ct.gov/resource"

    # create the first request
    req <-
      request(base) |>
      req_url_path_append(resource) |> #append the user provided 4x4 to the base url
      req_url_query(
        `$limit` = rows, # use the default rows=10000 or what ever the user provided
        `$offset` = 0 # start with the first row
      )

    # perform the request and then with each request increment the offset by `rows`
    resps <-
      req_perform_iterative(
        req,
        progress = FALSE,
        next_req = iterate_with_offset(
          param_name = "$offset",
          start = 0,
          offset = rows # use the default rows=10000 or what ever the user provided
        )
      )

    # convert the response objects from the API to a tibble of data
    # for each response from the API grab the json body, convert to an R list
    # and then squash all that into a singular tibble
    data <-
      resps |>
      resps_data(
        \(resp) {
          resp |>
            resp_body_json() |>
            bind_rows()
        }
      )

    return(data)
  }

# === main ===
main <- function() {
  # main entrypoint into our analysis

  # data ingestion
  data <- fetch_all_data(resource = RESOURCE)

  # key measures
  total_tests <-
    data |>
    mutate(date = ymd_hms(lastupdatedate)) |>
    filter(date == max(date)) |>
    reframe(
      total_pandemic_tests = sum(
        as.numeric(numberoftests),
        na.rm = TRUE
      )
    ) |>
    pull(total_pandemic_tests)

  cat(
    paste(
      "The total number of pandemic tests in these data were:",
      format(total_tests, big.mark = ',')
    )
  )
}

# === entrypoint ===
main()
```

One constant, one fetch function, one `main()` - a script you can hand
to someone else, and the only thing they need to know to run it is
that last line.
