---
title: "Databases, with DuckDB Standing In"
packages: [duckdb, DBI, dbplyr]
---

```{r}
#| label: chapter-root
#| include: false
library(dplyr)
library(readr)
library(arrow)
library(fs)
library(lubridate)
library(stringr)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

```{r}
#| label: rebuild-dataset
#| include: false
# Rebuild the project chapter's parquet dataset so this chapter stands alone.
LAB_DATASET <- path(tempdir(), "lab-results")

read_book_data(
  "synthetic-covid-lab-results.csv",
  col_types = cols(
    result_snomed = col_character(),
    specimen_source_snomed = col_character(),
    ordering_provider_npi = col_character(),
    patient_zip = col_character(),
    .default = col_guess()
  )
) |>
  mutate(
    performing_lab = str_to_title(performing_lab),
    collection_year = year(specimen_collection_dt),
    collection_month = month(specimen_collection_dt)
  ) |>
  write_dataset(
    LAB_DATASET,
    format = "parquet",
    partitioning = c("collection_year", "collection_month")
  )
```

The last chapter left a loose thread. To get one row per specimen, it did
this:

```r
lab_ds |>
  select(specimen_id, result_dt, result_snomed) |>
  collect() |>
  slice_max(result_dt, n = 1, by = specimen_id, with_ties = FALSE)
```

Look at where `collect()` sits. Three columns for **every row in the
feed** come into memory, and only then does the deduplication happen.
At 25,000 rows nobody notices. At sixteen million, that's the exact
mistake the arrow chapter warned about — and it isn't carelessness. The
query needs to rank rows within each specimen, and arrow's engine has no
window functions. There was nowhere else to put the `collect()`.

That's the wall for this part: not the size of the data, but the *kind
of question*. Which means it's time for a database.

## What a database gives you

This chapter uses DuckDB, but the point isn't DuckDB. It's that at some
size and some complexity, the answer stops being "a better file" and
becomes "a system that manages the data for you." What you get:

- **A query planner.** You describe the result you want; the engine
  decides the order of operations, which indexes to use, what to do in
  parallel. You stop hand-optimizing the order of your `filter()` calls.
- **SQL.** The most widely spoken language in data work, and the one
  every agency database already speaks.
- **Execution beyond memory.** A good engine spills to disk when a join
  or a sort outgrows RAM, instead of failing.
- **Persistence with real types and constraints.** A column is an
  integer because the schema says so, and it can be made impossible for
  a column to be null or a key to be duplicated.
- **Concurrency and transactions.** Several people query at once; a
  half-finished write never becomes half-visible.

### Two kinds of database, and which one you're in

Worth knowing, because the words come up in every conversation with a
database administrator:

**OLTP** — online *transaction* processing. Postgres, SQL Server, MySQL.
Built for many small reads and writes hitting a few rows at a time:
register a patient, record a result, update an address. Almost certainly
what your agency's surveillance system runs on.

**OLAP** — online *analytical* processing. Built for the opposite: a few
enormous reads scanning millions of rows to produce one summary table.
Columnar, like parquet, and for the same reasons.

You do OLAP work. The system of record is OLTP. That mismatch is why
"just run the query in production" is a request the DBA will refuse —
your scan of sixteen million rows would sit on top of the tables
clinicians are trying to write to.

## DuckDB: the analytical database with no server

DuckDB is an OLAP database that runs **inside your R session**. No
server, no port, no service account, no ticket. If SQLite is "a database
in a file," DuckDB is that for analytics.

```{r}
#| label: install
#| eval: false
pak::pak("duckdb")
```

```{r}
#| label: connect
library(DBI)
library(duckdb)

con <- dbConnect(duckdb(), dbdir = path(tempdir(), "lab.duckdb"))

dbGetQuery(con, "SELECT version() AS duckdb_version")
```

`dbConnect()` and `DBI` are not DuckDB-specific — that's the point. The
same two lines, pointed at a different driver, connect to Postgres or SQL
Server. Everything in this chapter except DuckDB's file-reading tricks is
knowledge that transfers directly to your agency's database.

::: {.callout-warning}
## Always close the connection

`dbDisconnect(con, shutdown = TRUE)` at the end. Without `shutdown =
TRUE`, DuckDB leaves its database process running and the file locked,
which is a confusing thing to debug an hour later.
:::

## Trick one: query files without importing them

Most databases make you load data in before you can query it. DuckDB
reads parquet and CSV files *as if they were tables*:

```{r}
#| label: query-parquet
lab_glob <- path(LAB_DATASET, "**", "*.parquet")

dbGetQuery(
  con,
  sprintf(
    "SELECT count(*) AS results,
            count(DISTINCT specimen_id) AS specimens
     FROM read_parquet('%s', hive_partitioning = true)",
    lab_glob
  )
)
```

Nothing was imported. DuckDB opened the parquet files the last chapter
wrote, read the columns the query mentioned, and answered. Hive
partitions come through as columns, exactly as they did in arrow:

```{r}
#| label: partitions
dbGetQuery(
  con,
  sprintf(
    "SELECT collection_year, count(*) AS n
     FROM read_parquet('%s', hive_partitioning = true)
     GROUP BY collection_year
     ORDER BY collection_year",
    lab_glob
  )
)
```

Note the `NA` row. Those are the specimens with no collection date —
still there, still counted, sitting in the partition named for the null
value. A database will not hide them from you, but it won't point at
them either.

Typing that `read_parquet(...)` every time gets old. Make it a **view** —
a saved query that behaves like a table:

```{r}
#| label: create-view
dbExecute(
  con,
  sprintf(
    "CREATE OR REPLACE VIEW lab_results AS
     SELECT * FROM read_parquet('%s', hive_partitioning = true)",
    lab_glob
  )
)

dbListTables(con)
```

The data still lives in the parquet files. The database file holds only
the definition — a few hundred bytes saying "when someone asks for
`lab_results`, go read those." Your dataset stays portable, and you get
to query it by name.

## Trick two: the same dplyr you already write

You do not have to write SQL. `dbplyr` translates `dplyr` into whatever
dialect the database speaks:

```{r}
#| label: dbplyr-basic
lab_results <- tbl(con, "lab_results")

lab_results |>
  group_by(performing_lab) |>
  summarise(tests = n(), .groups = "drop") |>
  arrange(desc(tests)) |>
  collect()
```

Identical to the arrow version, identical to plain `dplyr`, and lazy in
the same way — nothing runs until `collect()`.

`show_query()` is the part to linger on, because it's the best SQL
teacher you will meet:

```{r}
#| label: show-query
lab_results |>
  group_by(performing_lab) |>
  summarise(tests = n(), .groups = "drop") |>
  arrange(desc(tests)) |>
  show_query()
```

Write the `dplyr` you know, read the SQL it produces, and after a few
weeks of that you can write the SQL yourself. For anyone who arrived here
from SAS, this is the same trick that made `PROC SQL` learnable: see the
same question expressed two ways.

::: {.callout-tip}
## This is the transferable skill

`tbl(con, "table_name") |> dplyr verbs |> collect()` works against
Postgres, SQL Server, Snowflake, BigQuery, Redshift, and Spark. Change
the connection, keep the pipeline. When someone finally grants you
read access to the real surveillance database, you already know how to
query it — and `show_query()` is what you paste into the ticket when the
DBA asks what you intend to run.
:::

## Trick three: the query arrow couldn't do

Back to the dedupe. In SQL, "keep the latest result per specimen" is a
window function — number the rows within each specimen, newest first,
keep number one:

```{r}
#| label: dedupe-sql
dbGetQuery(
  con,
  "
  WITH ranked AS (
    SELECT
      specimen_id,
      result_snomed,
      row_number() OVER (
        PARTITION BY specimen_id
        ORDER BY result_dt DESC
      ) AS rn
    FROM lab_results
  )
  SELECT
    count(*) AS specimens,
    sum(CASE WHEN result_snomed IN ('260373001', '10828004') THEN 1 ELSE 0 END)
      AS positives
  FROM ranked
  WHERE rn = 1
  "
)
```

Or the same thing in `dplyr`, which `dbplyr` turns into that same SQL:

```{r}
#| label: dedupe-dplyr
lab_results |>
  group_by(specimen_id) |>
  dbplyr::window_order(desc(result_dt)) |>
  mutate(rn = row_number()) |>
  ungroup() |>
  filter(rn == 1) |>
  summarise(
    specimens = n(),
    positives = sum(
      ifelse(result_snomed %in% c("260373001", "10828004"), 1L, 0L),
      na.rm = TRUE
    )
  ) |>
  collect()
```

Now compare that positive count against the naive one — every row,
corrections and superseded results included:

```{r}
#| label: naive-count
lab_results |>
  summarise(
    all_rows = n(),
    positives_counting_every_row = sum(
      ifelse(result_snomed %in% c("260373001", "10828004"), 1L, 0L),
      na.rm = TRUE
    )
  ) |>
  collect()
```

**Forty-four positives that don't exist.** Small here. Scaled to a real
feed, that's the difference between a positivity rate you can defend and
one you can't — and no error was raised at any point along the way. The
whole computation stayed in the database; nothing but the two-row answer
crossed into R.

## Trick four: exact answers

The arrow chapter had to settle for an approximate median. DuckDB
doesn't:

```{r}
#| label: exact-median
dbGetQuery(
  con,
  "
  SELECT
    median(epoch(result_dt::TIMESTAMP) - epoch(specimen_collection_dt::TIMESTAMP))
      / 86400.0 AS median_days,
    quantile_cont(
      epoch(result_dt::TIMESTAMP) - epoch(specimen_collection_dt::TIMESTAMP),
      0.9
    ) / 86400.0 AS p90_days
  FROM lab_results
  WHERE specimen_collection_dt IS NOT NULL
  "
)
```

An exact median, and a 90th percentile — the number that actually
matters for turnaround, since the tail is what makes a count provisional.
Two notes on that query:

`epoch()` turns a timestamp into seconds, and the `::TIMESTAMP` casts
drop the time zone first. Subtracting two *time-zone-aware* timestamps
directly is an error in DuckDB, and `date_diff()` on them pulls in an
extension (see below). Casting sidesteps both.

::: {.callout-warning}
## Extensions and locked-down networks

DuckDB ships a small core and downloads optional **extensions** on
demand — time zones (`icu`), spatial, direct reads from S3 or HTTP. The
first query that needs one triggers a download, and on an agency network
that download may simply fail:

```
Extension Autoloading Error: An error occurred while trying to
automatically install the required extension 'icu'
```

That is a firewall, not a bug. Install what you need while you have
connectivity (`INSTALL icu;` once, then `LOAD icu;` per session), or
write queries that stay inside the core — which is what the casts above
are doing. Worth knowing *before* you're demoing to a room.
:::

## Trick five: it works with arrow, not against it

They are built by people who talk to each other. An arrow Dataset can be
handed to DuckDB without copying the data:

```{r}
#| label: to-duckdb
open_dataset(LAB_DATASET) |>
  to_duckdb() |>
  group_by(test_method) |>
  summarise(tests = n(), .groups = "drop") |>
  collect()
```

And results go back out to parquet, so the database is a step in the
pipeline rather than a place data goes to live:

```{r}
#| label: copy-out
deduped_path <- path(tempdir(), "deduped.parquet")

dbExecute(
  con,
  sprintf(
    "COPY (
       SELECT * EXCLUDE rn FROM (
         SELECT *, row_number() OVER (
           PARTITION BY specimen_id ORDER BY result_dt DESC
         ) AS rn
         FROM lab_results
       ) WHERE rn = 1
     ) TO '%s' (FORMAT PARQUET)",
    deduped_path
  )
)

file_size(deduped_path)
```

That's a deduplicated copy of the whole feed, written straight to
parquet, without a single row passing through R. (`SELECT * EXCLUDE rn`
is a DuckDB convenience — every column except that one. Standard SQL
makes you list all forty-three.)

You can also put R data *into* the database, which is how you join a
lookup table against the feed:

```{r}
#| label: write-table
dbWriteTable(
  con,
  "town_population",
  tibble(town = c("Hartford", "New Britain", "Avon"), population = c(121054, 74135, 18098)),
  overwrite = TRUE
)

dbGetQuery(con, "SELECT * FROM town_population")
```

Unlike the view, this one is really stored in the `.duckdb` file. Close
the connection, come back tomorrow, and it's there.

```{r}
#| label: disconnect
dbDisconnect(con, shutdown = TRUE)
```

## So which one do I use?

| Situation | Reach for |
|---|---|
| One file, fits in memory | `readr` and `dplyr`. Don't overthink it. |
| Bigger than memory, straightforward filters and group-bys | parquet + `arrow` |
| Window functions, multi-table joins, exact quantiles, SQL | DuckDB |
| Several people need the same data, or it must be written to | A server database — Postgres, SQL Server |
| It's the agency's system of record | That database, and a DBA to talk to |

These are not rivals. A normal workflow uses three of them: land the feed
as partitioned parquet, query it with arrow for the simple things, and
open a DuckDB connection when the question gets sharp. The files stay the
source of truth the whole time.

## A shout-out across the fence: polars

If you work anywhere near data engineers or Python-based modelers, you
will hear about **polars**. It's a dataframe library — closer in spirit
to `dplyr` than to a database — built on the same columnar ideas as
everything in this part: typed columns, lazy evaluation, a query
optimizer, execution that can exceed memory.

The reason it matters to you is not that you should switch. It's that
your parquet file is the handoff:

```python
import polars as pl

lab = pl.scan_parquet("data/processed/lab-results/**/*.parquet")

(
    lab.filter(pl.col("collection_year") == 2021)
    .group_by("test_method")
    .agg(pl.len().alias("tests"))
    .collect()
)
```

`scan_parquet` is lazy; `.collect()` runs it. If that reads like the
arrow chapter, that's the point — same three ideas, different keyboard.
Your colleague reads your dataset with their tools, with your types and
your column names intact, and neither of you emails a CSV. Python's own
`duckdb` package will run the identical SQL from this chapter, too.

Learn one of these properly and the others are dialects.

## A word about what this does not change

A `.duckdb` file is a database, not a vault. It is not encrypted, it does
not authenticate anyone, and copying it copies everything in it. Putting
identifiable data into an analytical database on your laptop is exactly
as consequential as putting it in a CSV on your laptop. The same data-use
agreement, the same storage rules, the same minimum-necessary principle.
Convenience is not permission.

## Where this leaves the part

Four chapters ago the problem was a file that wouldn't open. Now:

- **arrow** reads more data than fits, lazily, with `dplyr`.
- **parquet** stores it in a container that knows its own types and reads
  only what a query touches.
- **DuckDB** answers questions those two can't, in SQL that transfers to
  every database you'll ever be handed a connection string for.

None of it required a server, an administrator, or a budget line. That is
the actual lesson: the ceiling on what you can analyze is much higher
than "what fits in RAM," and getting above it is a matter of picking the
right container and the right engine.
