---
title: "Parquet: The Format Worth Knowing"
packages: [arrow]
---

```{r}
#| label: chapter-root
#| include: false
library(dplyr)
library(readr)
library(arrow)
library(fs)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Last chapter ended on an unfinished job. `open_dataset()` stopped R from
swallowing the whole file at once, but the file underneath was still a
CSV: still text to be parsed, still silent about its own types, still
unable to answer "how many rows?" without being read end to end.

This chapter swaps the container. The format is **Apache Parquet**, and
the argument of this chapter is that it is worth learning properly —
not as an arrow implementation detail, but as a thing you will keep
running into for the rest of your career, in and out of public health.

## What a parquet file actually is

Take five rows of the feed and think about how they sit on disk.

A CSV is **row-oriented**. Everything about test one, then everything
about test two:

```
NRL-20201029-03807,Nutmeg Regional...,2020-10-29T11:42:28Z,RT-PCR,260415000,Not detected
HVH21-004263,Housatonic Valley...,2021-04-03T12:38:05Z,Rapid immunoassay,260373001,Detected
```

Parquet is **column-oriented**. All the specimen IDs together, then all
the lab names together, then all the results together:

```
specimen_id      : NRL-20201029-03807 | HVH21-004263 | NRL-20210106-07234 | ...
performing_lab   : Nutmeg Regional... | Housatonic Valley... | Nutmeg Regional... | ...
result_snomed    : 260415000 | 260373001 | 260415000 | ...
```

That one rearrangement is where nearly everything else comes from. Four
consequences, in the order they will matter to you:

1. **Reading three columns costs three columns.** They're contiguous on
   disk. The other forty are never touched — not read, not decompressed,
   not skipped over byte by byte.
2. **Compression gets much better.** A column holds one kind of thing.
   Eight lab names repeating across sixteen million rows compress to
   almost nothing; the same eight names scattered through a row-oriented
   file do not.
3. **Types are stored, not guessed.** Each column carries its type in the
   file's footer. `result_snomed` is a string because the file says it's
   a string.
4. **The file describes itself.** Row counts, column names, types, and
   per-chunk minimum and maximum values all live in the footer, so a lot
   of questions get answered without reading any data at all.

## Writing one

`write_parquet()` takes a data frame and a path. That's the whole API.

```{r}
#| label: write-parquet
lab_results <-
  read_book_data(
    "synthetic-covid-lab-results.csv",
    col_types = cols(
      result_snomed = col_character(),
      specimen_source_snomed = col_character(),
      ordering_provider_npi = col_character(),
      patient_zip = col_character(),
      .default = col_guess()
    )
  )

parquet_path <- path(tempdir(), "lab-results.parquet")

write_parquet(lab_results, parquet_path)

file_size(parquet_path)
```

Compare that against the CSV it came from:

```{r}
#| label: size-comparison
csv_path <- book_data_path("synthetic-covid-lab-results.csv")

tibble(
  format = c("csv", "parquet"),
  size = as_fs_bytes(c(file_size(csv_path), file_size(parquet_path)))
) |>
  mutate(times_smaller = round(as.numeric(size[1]) / as.numeric(size), 1))
```

Roughly a sixth of the size, holding exactly the same 25,425 rows. No
sampling, no dropped columns, nothing thrown away — it's the same data in
a container that doesn't waste space.

## Types survive the round trip

This is the part that should sell you, and it has nothing to do with size
or speed.

```{r}
#| label: type-round-trip
from_parquet <- read_parquet(parquet_path)

tibble(
  column = c(
    "result_snomed",
    "ordering_provider_npi",
    "specimen_collection_dt",
    "ct_value"
  ),
  original = c(
    class(lab_results$result_snomed)[1],
    class(lab_results$ordering_provider_npi)[1],
    class(lab_results$specimen_collection_dt)[1],
    class(lab_results$ct_value)[1]
  ),
  after_parquet = c(
    class(from_parquet$result_snomed)[1],
    class(from_parquet$ordering_provider_npi)[1],
    class(from_parquet$specimen_collection_dt)[1],
    class(from_parquet$ct_value)[1]
  )
)
```

Identical. No `col_types` argument, no guessing, no chunk of the file
sampled to work it out. You fixed those types **once**, when you wrote
the file, and the fix is now a property of the file rather than a block
of code every reader has to remember to include.

Watch a CSV fail the same test:

```{r}
#| label: csv-round-trip
small <- tibble(
  n_tests = 1:5L,
  rate = c(1.5, 2, 3, 4, 5),
  collected = as.Date("2020-03-01") + 0:4,
  zip = c("06001", "06002", "06003", "06004", "06005")
)

csv_tmp <- path(tempdir(), "types.csv")
write_csv(small, csv_tmp)

tibble(
  column = names(small),
  original = sapply(small, \(x) class(x)[1]),
  after_csv = sapply(read_csv(csv_tmp, show_col_types = FALSE), \(x) class(x)[1]),
  after_parquet = {
    pq_tmp <- path(tempdir(), "types.parquet")
    write_parquet(small, pq_tmp)
    sapply(read_parquet(pq_tmp), \(x) class(x)[1])
  }
)
```

`readr` does well here — it recovers the date and keeps the ZIP as text,
because modern `readr` is good at guessing. But look at `n_tests`: it
went in as an **integer** and came back a **double**. The CSV had no way
to say "this is a count, not a measurement," so the information was
simply lost in transit. Parquet returns exactly what it was given.

::: {.callout-important}
## Why an epidemiologist should care about this specifically

The bugs this prevents are the ones that bite quietly.

- A ZIP code or FIPS code losing its leading zero, so `06001` joins
  against nothing and a town silently drops out of a map.
- A SNOMED or LOINC code read as a number, written back out in
  scientific notation, and never matching the code table again.
- A date column that parses as a date on your machine and as text on a
  colleague's, because a locale differs.
- An MRN that fits in a double until the day it doesn't.

None of these throws an error. They produce a number that is wrong, in a
report that looks fine. Storing types in the file removes the entire
class.
:::

## Why it's faster

Three mechanisms, and it's worth knowing which is which, because they
reward different habits.

**Column projection.** `select()` fewer columns and parquet reads less
disk, in direct proportion. On a forty-three column feed where your query
touches four, that is roughly a tenfold reduction in bytes read before
anything else happens.

**Row groups and statistics.** A parquet file is internally divided into
row groups — typically a few hundred thousand rows each — and the footer
records the minimum and maximum of every column in every row group. Ask
for `specimen_collection_dt >= "2021-01-01"` and the engine reads the
footer, sees that a row group tops out at 2020-06-30, and never opens it.
This is called **predicate pushdown**, and it's why filtering early
matters so much.

**No parsing.** A double is stored as a double. There is no text to
convert.

::: {.callout-note}
## Measured, not guessed

The numbers below come from the 2,034,000-row version of this same feed
(`just lab-data 2000000`), queried with a columnar engine on one laptop,
on a RAM-backed temporary filesystem. Absolute times on your machine will
differ and the ratios will move around. The *shape* is the point.

| File | Size |
|---|---|
| `lab-results.csv` | 1,310 MB |
| `lab-results.parquet` (snappy, the default) | 193 MB |
| `lab-results.parquet` (zstd) | 140 MB |
| partitioned by year and month (zstd, 19 files) | 136 MB |

| Query over 2.03M rows | From CSV | From parquet |
|---|---|---|
| monthly test counts and positives | 1.86 s | 0.01 s |
| count distinct performing labs | 1.80 s | < 0.01 s |
| count one month of collections | 1.73 s | 0.01 s |
| same count, partitioned by month | — | 0.004 s |

The CSV timings barely move between those three queries, because all
three do the same work: parse the entire file. That is the tell. With
parquet, the cost tracks what you actually asked for.
:::

## Datasets: many files, one table

One parquet file is good. A **dataset** — a directory of parquet files
that arrow treats as a single table — is where this starts paying for
itself on real surveillance data.

`write_dataset()` splits the data across files using columns you choose:

```{r}
#| label: write-dataset
lab_dir <- path(tempdir(), "lab-dataset")

lab_results |>
  mutate(
    collection_year = lubridate::year(specimen_collection_dt),
    collection_month = lubridate::month(specimen_collection_dt)
  ) |>
  write_dataset(
    path = lab_dir,
    format = "parquet",
    partitioning = c("collection_year", "collection_month")
  )

dir_tree(lab_dir, recurse = 1)
```

Those folder names are not decoration. `collection_year=2021` is a
convention called **Hive partitioning**, and every tool that reads
parquet understands it. When you open the directory, the partition
columns come back as real columns even though they aren't stored inside
any file — they're in the paths:

```{r}
#| label: read-dataset
lab_ds <- open_dataset(lab_dir)

lab_ds |>
  filter(collection_year == 2021, collection_month == 1) |>
  group_by(test_method) |>
  summarise(tests = n(), .groups = "drop") |>
  collect()
```

That query never opened a single 2020 file. Not "read them and discarded
them" — never opened them. On ten years of surveillance data partitioned
by year, a question about last year costs one year of reading.

::: {.callout-warning}
## Partition on what you filter on — and don't overdo it

Partitioning is the one setting here that people get badly wrong.

**Good partition keys** are low-cardinality columns that show up in your
`filter()` calls constantly: year, month, disease, jurisdiction,
reporting county.

**Bad partition keys** are high-cardinality: patient ID, specimen ID,
town-by-day. Partitioning by patient on sixteen million tests means
millions of tiny files, and reading a million 4 KB files is *slower* than
reading one big one — you've replaced a data problem with a filesystem
problem.

Aim for partitions that leave files somewhere in the tens-to-hundreds of
megabytes. If a partition would produce a 30 KB file, that partition is
too fine.

One more gotcha visible in the tree above: rows whose partition key is
`NA` land in a directory named for the null value rather than being
dropped. Missing collection dates don't vanish — they collect somewhere
you'll only find if you look.
:::

## Compression is a dial

Snappy is the default: fast to write, fast to read, decent ratio. When a
dataset is going to be written once and read for years — an annual
archive, say — zstd is usually the better trade: a noticeably smaller
file, slightly more CPU on write, and read speed close enough that you
won't notice.

There's a catch worth meeting now rather than in production. The `arrow`
R package is a wrapper around a C++ library, and **which codecs are
available depends on how that library was built** on your machine. Ask
before you assume:

```{r}
#| label: codecs
sapply(c("snappy", "gzip", "zstd", "lz4"), codec_is_available)
```

If `zstd` is `TRUE` there, use it for archival datasets:

```{r}
#| label: compression
if (codec_is_available("zstd")) {
  zstd_path <- path(tempdir(), "lab-results-zstd.parquet")
  write_parquet(lab_results, zstd_path, compression = "zstd")

  tibble(
    file = c("snappy (default)", "zstd"),
    size = as_fs_bytes(c(file_size(parquet_path), file_size(zstd_path)))
  )
} else {
  "zstd not available in this arrow build - snappy it is."
}
```

If it's `FALSE`, you have a partial build. On Linux that usually means
`arrow` compiled against a system library without the optional codecs;
reinstalling with `Sys.setenv(NOT_CRAN = "true")` set before
`pak::pak("arrow")` generally pulls a full-featured build. It is worth
sorting out — `arrow_info()` will show you the whole capability list,
including whether S3 reading is compiled in — but nothing else in this
chapter depends on it. A parquet file written with snappy is still a
parquet file.

## Why this is worth knowing beyond this book

Everything so far has been about one feed on one laptop. The reason to
learn parquet properly is that it isn't an R thing, and it isn't an
epidemiology thing.

**It is the interchange format between languages.** A colleague working
in Python reads your parquet file with `pandas.read_parquet()` or
`polars.read_parquet()` and gets your exact types back — no CSV
round-trip, no "why is the date a string on my end," no argument about
encodings. The file *is* the contract.

**It is what the platforms underneath you already use.** [DuckDB](duckdb.qmd), Spark,
Databricks, BigQuery, Snowflake, Amazon Athena, and most cloud data lakes
either store parquet natively or read it directly. If your agency ever
moves its surveillance data to a cloud warehouse, it is landing in
parquet whether or not anyone tells the epidemiologists.

**It reads well over a network.** Because the footer describes the file,
a tool can fetch just the footer, work out which byte ranges it needs,
and pull only those — from S3 or Azure Blob storage, without downloading
the file. You cannot do that with a CSV.

**It is a good archival format.** Self-describing, compressed, and a
stable open specification with multiple independent implementations. A
parquet file written today opens in ten years without the sidecar
document explaining what column 17 meant.

For public health work specifically, the ones that come up most:

- **Annual surveillance archives.** Partition by year, keep a decade
  online, query any slice without a database server.
- **Line lists that outgrow Excel.** Which is most of them, eventually.
- **Handing data to a modeler.** They get your types, your missingness,
  and your column names exactly as you meant them.
- **Reproducibility.** The types are in the file, so a re-run in two
  years reads the same data the same way. No `col_types` block to
  maintain, and no chance of a colleague forgetting it.

## When *not* to use parquet

Being fair about it:

- **Someone needs to open it in Excel.** Excel doesn't read parquet.
  Export a CSV or an `.xlsx` for that person and keep parquet for the
  pipeline.
- **The data is small.** A 200-row lookup table is fine as a CSV. Use
  the thing you can read with `less`.
- **You want it in version control with readable diffs.** Parquet is
  binary; git can store it but can't show you what changed.
- **You're appending one row at a time.** Parquet is written in
  column-chunk batches, not appended to row by row. Log to something
  else, convert on a schedule.
- **It is not encryption.** Parquet is compressed and binary, which makes
  it unreadable in a text editor and *not at all* protected. Sensitive
  data needs the same access controls, storage rules, and agreements it
  always did. Nothing in this chapter changes your data-use agreement.

## Where this leaves the brief

You now have a container that stores its own types, compresses to a
fraction of the CSV, and reads only what a query asks for. The project
chapter puts it to work: convert the feed once, partition it sensibly,
and answer both of the Commissioner's questions from it.
