---
title: "Cleaning Data"
packages: [janitor, tidyr, stringr, lubridate]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Real data is rarely clean. Column names are inconsistent, values need
reshaping, text fields need tidying up, and dates need parsing before
they're ready for analysis. This chapter covers four packages that
handle the bulk of that work.

## Installing Packages

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "janitor", # for cleaning names and basic data cleaning
  "tidyr", # for reshaping data
  "stringr", # for working with text
  "lubridate" # for working with dates
))
```

```{r}
#| label: setup-libraries

library(dplyr)
library(readr)
library(here)
library(janitor)
library(tidyr)
library(stringr)
library(lubridate)
```

```{r}
#| label: load-data
#| include: false

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```


## Common Data Cleaning Patterns

Data come in all shapes and sizes, and it is difficult to come up with a
single approach that will fit every situation. However, there are some common
patterns that show up frequently in public health data. This chapter will cover 
some of the most common data cleaning tasks and the R packages that can help 
you accomplish them.


```{r}
#| label: loading-data
#| eval: false

# load up our raw data
csv_path <- here("data", "raw", "ct_covid_by_town_2026-06-30.csv")
covid_data <- read_csv(csv_path)
```

### Working with Names

By now users may have noticed that the column names in our `covid_data` 
are not very easy to work with. They contain spaces and mixed case that
are difficult to reference due to R's case-sensitive nature.  Other data
may have special characters, punctuation or exta white space in the names of
their columns. 

The `janitor` package has a function called `clean_names()` 
that can help with this. It will convert column names to lower case, replace 
spaces with underscores, and remove special characters. This makes it much 
easier to work with the data in R. Users can easily recreate this
functionality from scratch, but `clean_names()` is a convenient and 
well-tested function that saves time and effort and is such an easy package
to recommend.


Recall the names of our `covid_data`: 

```{r}
#subset of the names
names(covid_data) |> head(5)
```

```{r}
#| label: clean-names-janitor

covid_data_clean <-
  covid_data |>
  clean_names()

names(covid_data_clean) |> head(5)
```

A single function call - it is that simple.  `janitor` package has
a lot of other useful functions for cleaning, and even presenting data in a 
more readable format.  We'll cover some of those later, but I encourage users
to explore the package documentation in the package callout.


### Working with Dates

Earlier in the book we covered `readr`'s parsing functions for reading in 
dates. `readr` is great, but there is a package whose only job is to
work with dates and it makes incredibly ergonomical and intuitive.

Users may have noticed that the `last_update_date` column in our 
`covid_data_clean` is not considered a date in R, but rather a 
character string. If users are not going to be manipulating the dates, 
this is not a problem. However, if users want to do any date math, such as 
calculating the number of days between two dates, or filtering for dates 
within a certain range, they will need to convert the
character string to a date object. 

Take a look at `last_update_date`:

```{r}
covid_data_clean |>
  select(last_update_date) |> # select ONLY this column
  slice(1:5) # slice the first 5 rows - kind of like head()
```

I don't know about you, but I can never remember the date format jargon that
is common in many date parsing functions, and with lubridate you don't have 
to.  Those dates look like the month, then the day and then the year.


We will tell lubridate that the date is in the month-day-year format by 
using its `mdy()` function and it handles the rest.

```{r}

# we are going to change the character string to a date and then save the changes
# back to itself

covid_data_clean <-
  covid_data_clean |>
  mutate(last_update_date = mdy(last_update_date)) # mutate for manipulating columns

covid_data_clean |>
  select(last_update_date) |>
  slice(1:5)
```


Did you catch the `<date>` below the column name in the output? It is a date 
now and can be seen in R's standard date format.  We can now do date 
operations on this column.

We can do simple things like this...

```{r}
# how many days are in the dataset?
max(covid_data_clean$last_update_date) - min(covid_data_clean$last_update_date)
```

or more complex things like this...

```{r}
# average total deaths per calendar week - demonstration purposes only

covid_data_clean |>
  group_by(week = floor_date(last_update_date, "week")) |> # here we are using lubridate's floor_date() to round down to the nearest week
  summarise(avg_total_deaths = mean(total_deaths, na.rm = TRUE)) |> # here we are using dplyr's summarise() to calculate the average total deaths per week
  slice(1:5)
```

It is important to keep in mind what folks want to do with these data
and then clean them in a manner that will make it easier to do those things.  




TODO: clean_names(), tabyl(), get_dupes(), other janitor basics.

## `tidyr`

TODO: pivot_longer(), pivot_wider(), drop_na()/replace_na(), separate/unite.

## `stringr`

TODO: str_detect(), str_replace(), str_trim(), str_to_title() and friends.

## `lubridate`

TODO: ymd(), as_date(), wday(), floor_date() and other date parsing basics.
