---
title: "Cleaning Data"
packages: [janitor, tidyr, stringr]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
source(here::here("preamble-packages.R"))
.chapter_data <- here::here("chapters", "01-working-with-real-data", "data")
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Real data is rarely clean. Column names are inconsistent, values need
reshaping, and text fields need tidying up before they're ready for
analysis. This chapter covers three packages that handle the bulk of
that work.

## Installing Packages

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "janitor", # for cleaning names and basic data cleaning
  "tidyr", # for reshaping data
  "stringr" # for working with text
))
```

```{r}
#| label: setup-libraries

library(dplyr)
library(readr)
library(here)
library(janitor)
library(tidyr)
library(stringr)
```

```{r}
#| label: load-data
#| include: false

csv_path <- fs::path(.chapter_data, "ct-covid-by-town-2026-06-30.csv")
covid_data <- read_csv(csv_path)
```

## `janitor`

TODO: clean_names(), tabyl(), get_dupes(), other janitor basics.

## `tidyr`

TODO: pivot_longer(), pivot_wider(), drop_na()/replace_na(), separate/unite.

## `stringr`

TODO: str_detect(), str_replace(), str_trim(), str_to_title() and friends.
