---
title: "Cleaning Data"
packages: [janitor, tidyr, stringr, lubridate, glue]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Real data is rarely clean. Column names are inconsistent, values need
reshaping, text fields need tidying up, and dates need parsing before
they're ready for analysis. This chapter covers four packages that
handle the bulk of that work.

## Installing Packages

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "janitor", # for cleaning names and basic data cleaning
  "tidyr", # for reshaping data
  "stringr", # for working with text
  "glue", # for string interpolation
  "lubridate" # for working with dates
))
```

```{r}
#| label: setup-libraries

library(dplyr)
library(readr)
library(here)
library(janitor)
library(tidyr)
library(stringr)
library(lubridate)
library(glue)
```

```{r}
#| label: load-data
#| include: false

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```


## Common Data Cleaning Patterns

Data come in all shapes and sizes, and it is difficult to come up with a
single approach that will fit every situation. However, there are some common
patterns that show up frequently in public health data. This chapter will cover 
some of the most common data cleaning tasks and the R packages that can help 
you accomplish them.


```{r}
#| label: loading-data
#| eval: false

# load up our raw data
csv_path <- here("data", "raw", "ct_covid_by_town_2026-06-30.csv")
covid_data <- read_csv(csv_path)
```

### Working with Names

By now you've probably noticed that the column names in our `covid_data`
aren't very easy to work with. They contain spaces and mixed case that
are difficult to reference due to R's case-sensitive nature. Other data
may have special characters, punctuation, or extra white space in the
names of their columns.

The `janitor` package has a function called `clean_names()`
that can help with this. It will convert column names to lower case, replace
spaces with underscores, and remove special characters. This makes it much
easier to work with the data in R. You could recreate this functionality
from scratch, but `clean_names()` is a convenient and well-tested function
that saves time and effort and is such an easy package to recommend.


Recall the names of our `covid_data`: 

```{r}
#subset of the names
names(covid_data) |> head(5)
```

```{r}
#| label: clean-names-janitor

covid_clean <-
  covid_data |>
  clean_names()

names(covid_clean) |> head(5)
```

A single function call - it is that simple. The `janitor` package has
a lot of other useful functions for cleaning, and even presenting data in a
more readable format. We'll cover some of those later, but it's worth
exploring the package documentation in the callout above too.


### Working with Dates

Earlier in the book we covered `readr`'s parsing functions for reading in 
dates. `readr` is great, but there is a package whose only job is to
work with dates and it makes incredibly ergonomical and intuitive.

You may have noticed that the `last_update_date` column in our
`covid_clean` isn't considered a date in R, but rather a
character string. If you're not going to be manipulating the dates,
this isn't a problem. However, if you want to do any date math - such as
calculating the number of days between two dates, or filtering for dates
within a certain range - you'll need to convert the
character string to a date object.

Take a look at `last_update_date`:

```{r}
covid_clean |>
  select(last_update_date) |> # select ONLY this column
  slice(1:5) # slice the first 5 rows - kind of like head()
```

Date format jargon is easy to lose track of, and with `lubridate` you
don't have to remember it. Those dates look like the month, then the
day, then the year.


We will tell lubridate that the date is in the month-day-year format by 
using its `mdy()` function and it handles the rest.

```{r}

# we are going to change the character string to a date and then save the changes
# back to itself

covid_clean <-
  covid_clean |>
  mutate(last_update_date = mdy(last_update_date)) # mutate for manipulating columns

covid_clean |>
  select(last_update_date) |>
  slice(1:5)
```


Did you catch the `<date>` below the column name in the output? It is a date 
now and can be seen in R's standard date format.  We can now do date 
operations on this column.

We can do simple things like this...

```{r}
# how many days are in the dataset?
max(covid_clean$last_update_date) - min(covid_clean$last_update_date)
```

or more complex things like this...

```{r}
# average total deaths per calendar week - demonstration purposes only

covid_clean |>
  group_by(week = floor_date(last_update_date, "week")) |> # here we are using lubridate's floor_date() to round down to the nearest week
  summarise(avg_total_deaths = mean(total_deaths, na.rm = TRUE)) |> # here we are using dplyr's summarise() to calculate the average total deaths per week
  slice(1:5)
```

It is important to keep in mind what folks want to do with these data
and then clean them in a manner that will make it easier to do those things.

## `tidyr`

`tidyr` is a wonderful package for creating and reshaping [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html).

Pivoting data is a common usecase for `tidyr`. Our `covid_clean` dataset
as-is, is probably fine for most usecases.  However, it was not uncommon for 
similar data to be presented in a wider format during the pandemic. Let's
purposely un-tidy our data by making it wide and use the town names 
as the column names, something that should be a value in the data. We will
then take our wide 'messy' data and make it tidy again in a longer format.

### Wider

`pivot_wider()` is a function that takes a longer dataset and makes it wider.  

```{r}
wide <-
  covid_clean |>
  pivot_wider(
    id_cols = last_update_date, # this helps uniquely identifies each row in the dataset
    names_from = town, # think to yourself, "what do I want to use as the new column names?"
    values_from = total_cases, # think to yourself, "what do I want to use as the values in the new columns?"
    names_prefix = "cases_"
  ) |>
  clean_names()

wide |>
  head()
```

Goodness -  that is a lot of columns!

Now let's take that wide dataset and make it long again.

```{r}
long <-
  wide |>
  pivot_longer(
    cols = !last_update_date, # pivot everyhing BUT last_update_date.  Folks can all use a helper: starts_with("cases_")
    names_to = "town", # this is the name of the new column that will contain the town names
    values_to = "total_cases" # this is the name of the new column that will contain the total cases
  )

long |>
  head()
```



### Handling missing values with `drop_na()` and `replace_na()`

Not every column was filled in from day one. Early in the pandemic, CT
DPH hadn't started splitting cases into confirmed vs. probable yet -
`confirmed_cases` and `probable_cases` are `NA` for the earliest
reports.

```{r}
covid_clean |>
  summarise(missing_confirmed = sum(is.na(confirmed_cases)))
```

That's a real gap, not a typo, and you have two honest ways to handle
it depending on what comes next.

`drop_na()` removes rows missing a given column - use it when an
analysis genuinely needs the confirmed/probable breakdown and rows
without it can't contribute:

```{r}
covid_clean |>
  drop_na(confirmed_cases) |>
  nrow()
```

`replace_na()` keeps every row and fills the gap with a value you
choose:

```{r}
covid_clean |>
  replace_na(list(confirmed_cases = 0, probable_cases = 0)) |>
  select(last_update_date, confirmed_cases, probable_cases) |>
  head()
```

::: {.callout-warning}
Filling `NA` with `0` says "this was measured, and the answer was
zero." That's not what happened here - the confirmed/probable split
simply wasn't being tracked yet. Replacing with `0` would make early
2020 look like it had zero confirmed cases, which is misleading in a
public health report. `drop_na()` or leaving the `NA` in place is
usually the more honest choice; only reach for `replace_na()` when
missing genuinely does mean the same thing as zero (or whatever value
you're filling in with).
:::

### Splitting and combining columns with `separate()` and `unite()`

Sometimes several pieces of information need to travel together in one
column - a "2026-06" style label for grouping by month, for example.
`unite()` glues columns together with a separator:

```{r}
dates_split <-
  covid_clean |>
  mutate(
    report_year = year(last_update_date),
    report_month = month(last_update_date)
  ) |>
  select(last_update_date, report_year, report_month) |>
  distinct()

dates_united <-
  dates_split |>
  unite(col = "year_month", report_year, report_month, sep = "-", remove = FALSE)

dates_united |>
  head()
```

`separate()` does the reverse - split one column into several using a
separator to know where to break it apart:

```{r}
dates_united |>
  select(last_update_date, year_month) |>
  separate(col = year_month, into = c("year", "month"), sep = "-") |>
  head()
```

Just like the wide/long round trip earlier, `unite()` and `separate()`
are inverses of each other - whichever shape a report or a join needs,
the other function gets you back.

## `stringr` and `glue`

Public health data is full of free text - town names, facility names,
notes fields, category labels typed by different people at different
times. `stringr` gives you consistent functions for detecting,
replacing, and reshaping that text. `glue` turns data into sentences,
useful for report text, labels, and messages that need to fill in with
real values from your dataset instead of being copy-pasted by hand.

### Detecting patterns with `str_detect()`

`str_detect()` returns `TRUE`/`FALSE` for whether a pattern appears in
each string. Use it inside `filter()` to keep only rows that match.

```{r}
#| label: str-detect
covid_clean |>
  distinct(town) |>
  filter(str_detect(town, "New"))
```

That caught more than towns actually named "New Something" -
`Newington` and `Newtown` matched too, because `str_detect()` looks for
the pattern anywhere in the string. Anchor the pattern to the start
with `^` and require a space after "New" to match only the compound
names:

```{r}
#| label: str-detect-anchored
covid_clean |>
  distinct(town) |>
  filter(str_detect(town, "^New "))
```

### Changing case with `str_to_title()` and friends

Data entered by hand rarely agrees on case. `str_to_upper()`,
`str_to_lower()`, and `str_to_title()` normalize it:

```{r}
#| label: str-case
"HARTFORD" |> str_to_title()
"hartford" |> str_to_upper()
"Hartford" |> str_to_lower()
```

Reach for `str_to_title()` when preparing labels for a plot or report
where you want consistent, presentable capitalization no matter how the
value happened to be stored.

### Trimming whitespace with `str_trim()`

Stray leading or trailing whitespace is invisible in a viewer but
breaks joins and comparisons. `str_trim()` removes it:

```{r}
#| label: str-trim
"  Hartford   " |> str_trim()
```

If a `filter()` or a join isn't matching rows you know should match,
whitespace like this is one of the first things worth checking.

### Replacing values with `str_replace()`

`str_replace()` swaps the first match of a pattern; `str_replace_all()`
swaps every match. Both take the same three arguments: the string, the
pattern to find, and the replacement.

```{r}
#| label: str-replace
covid_clean |>
  distinct(town) |>
  filter(str_detect(town, "^New ")) |>
  mutate(short_name = str_replace(town, "^New ", "N. "))
```

### Building sentences with `glue()`

`glue()` interpolates R values directly into a string. Wrap any
expression in `{}` and it is evaluated and inserted in place:

```{r}
#| label: glue-basic
town_name <- "Hartford"

case_count <-
  covid_clean |>
  filter(town == town_name, last_update_date == max(last_update_date)) |>
  pull(total_cases)

glue("As of the latest report, {town_name} had {case_count} total cases.")
```

`glue()` also works row-by-row against a data frame column, which is
the more common way you'll use it in practice:

```{r}
#| label: glue-vectorized
covid_clean |>
  filter(last_update_date == max(last_update_date)) |>
  slice(1:3) |>
  mutate(
    summary = glue("{town}: {total_cases} total cases, {total_deaths} deaths")
  ) |>
  pull(summary)
```

This is the kind of line that goes straight into an automated report -
no manually copying numbers into a sentence, and no risk of a stale
number left over from last week's version.

## `lubridate`

Earlier in this chapter you used `mdy()` to parse `last_update_date`
and `floor_date()` to group rows by week. `lubridate` has a parsing
function for every common date order - the function name spells out
the order of year (`y`), month (`m`), and day (`d`) that it expects in
the string:

```{r}
#| label: lubridate-family
ymd("2026-06-30")
mdy("06/30/2026")
dmy("30-06-2026")
```

Pick whichever one matches the order used in your source data - no
format string, no `strptime` codes to look up.

### Converting to dates with `as_date()`

Sometimes a column is already a date-*time* (`POSIXct`), with hours,
minutes, and seconds attached, but the analysis only needs the
calendar date. `as_date()` truncates it:

```{r}
#| label: as-date
now() # a datetime, with a time attached
now() |> as_date() # a calendar date, time dropped
```

### Pulling out the day of the week with `wday()`

CT DPH published new case counts on a schedule, and gaps or slowdowns
in reporting often line up with weekends. `wday()` extracts the day of
the week from a date, and `label = TRUE` gives you a readable name
instead of a number:

```{r}
#| label: wday-demo
covid_clean |>
  distinct(last_update_date) |>
  mutate(day_of_week = wday(last_update_date, label = TRUE)) |>
  count(day_of_week)
```

That count is a quick way to check whether report dates cluster on
certain days of the week - useful for spotting reporting cadence, or
gaps, before you build anything downstream that assumes daily data.

::: {.callout-tip}
## The `lubridate` family, at a glance

| Function | What it does |
|---|---|
| `ymd()`, `mdy()`, `dmy()` | Parse a character string into a date, given the order of year/month/day |
| `as_date()` | Convert a datetime (`POSIXct`) down to a calendar date |
| `wday(x, label = TRUE)` | Extract the day of the week as a name |
| `floor_date(x, "week")` | Round a date down to the start of its week (covered earlier in this chapter) |
:::
