---
title: "Cleaning the Full Report"
---

```{r}
#| label: project-root
#| include: false
library(here)
source(here::here("preamble-data.R"))
```

The 4 p.m. call didn't stay about Hartford. Someone asked about
Stamford, and you didn't have it — you'd cleaned exactly one town, by
hand. The Commissioner's follow-up is blunt: **stop picking towns one at
a time. Clean the whole report once, so any town's number is ready
before the question gets asked.**

Same brief, wider net: this week's Situation Brief still leads with
Hartford, but you never want to be caught flat again when the call turns
to some other town.

## Setup

```{r}
#| eval: false

library(dplyr)
library(readr)
library(here)
library(janitor)
library(tidyr)
library(stringr)
library(lubridate)
library(glue)

covid_data <- read_csv(here("data", "raw", "ct_covid_by_town_2026-06-30.csv"))
```

```{r}
#| label: setup-exec
#| include: false

library(dplyr)
library(readr)
library(here)
library(janitor)
library(tidyr)
library(stringr)
library(lubridate)
library(glue)

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```

## Step 1 — Clean every column name at once

Chapter 1 renamed a handful of columns by hand with `select()`. That
works for one town and four columns; it does not scale to every column
in the dataset. `clean_names()` fixes all of them in a single call.

```{r}
covid_clean <-
  covid_data |>
  clean_names()

names(covid_clean)
```

## Step 2 — Parse the date once, for every row

```{r}
covid_clean <-
  covid_clean |>
  mutate(last_update_date = mdy(last_update_date))

covid_clean |>
  select(last_update_date) |>
  slice(1:3)
```

## Step 3 — Guard the town names with `stringr`

The town names in this file are already tidy, but a report that gets
re-downloaded every week shouldn't depend on that staying true.
Trimming whitespace and standardizing case costs nothing when the data
is already clean, and saves a debugging session the week it isn't.

```{r}
covid_clean <-
  covid_clean |>
  mutate(town = town |> str_trim() |> str_to_title())
```

## Step 4 — Daily new cases, for every town

Chapter 1's caveat about "as reported" data applies here at full scale.
`total_cases` is cumulative, so a daily new-case number means computing
`total_cases - lag(total_cases)` again — but this time across 169
towns at once, which means grouping by `town` first. Without the
`group_by()`, `lag()` would subtract the last row of one town from the
first row of the next.

```{r}
covid_clean <-
  covid_clean |>
  arrange(town, last_update_date) |>
  group_by(town) |>
  mutate(new_cases = total_cases - lag(total_cases)) |>
  ungroup()
```

Chapter 1 warned this could happen; at 169 towns, now it actually does.
Go look:

```{r}
#| label: find-negatives
covid_clean |>
  filter(new_cases < 0) |>
  select(town, last_update_date, new_cases) |>
  arrange(new_cases) |>
  head()
```

Those negative "new case" counts are real rows in your own output, not a
bug you introduced. Each one is a day a town's cumulative total came in
*lower* than its previous report — a correction, a de-duplicated test, a
late reversal. Subtracting two "as of" snapshots faithfully records that
as negative growth.

So what do you do with them? **Not** silently floor them to zero — that
hides a real data-quality signal, and the brief is supposed to be honest.
You leave them in and you don't let a single day drive a conclusion. That
is exactly why every verdict in this book averages over a week: one
corrected report can't flip the answer when it's one of seven. Knowing
these rows exist is the fix; the weekly average is the guard rail.

## Step 5 — Scope to the latest report

```{r}
latest_date <- max(covid_clean$last_update_date)

this_week <-
  covid_clean |>
  filter(last_update_date == latest_date)

nrow(this_week)
```

## Step 6 — Generate a one-line summary per town with `glue`

This is the payoff, and the answer to the ask that opened the chapter.
The next time the call turns to Stamford — or any of the other 168 towns
— the sentence is already sitting there, no filter to re-run under
pressure.

```{r}
town_lines <-
  this_week |>
  mutate(
    line = glue(
      "{town}: {total_cases} total cases ({new_cases} new), {total_deaths} deaths"
    )
  ) |>
  select(town, line)

town_lines |>
  filter(town %in% c("Hartford", "Stamford", "New Haven")) |>
  pull(line)
```

## Step 7 — A weekly trend appendix with `tidyr`

The Commissioner also wants to see the trend, not just the latest
snapshot, for the state's largest cities. `floor_date()` groups rows
into weeks, and `pivot_wider()` lays those weeks out as columns — a
compact table that reads left-to-right like a trend line.

```{r}
cities_of_interest <- c(
  "Hartford",
  "New Haven",
  "Bridgeport",
  "Stamford",
  "Waterbury"
)

weekly_trend <-
  covid_clean |>
  filter(town %in% cities_of_interest) |>
  mutate(week = floor_date(last_update_date, "week")) |>
  group_by(town, week) |>
  summarise(avg_new_cases = mean(new_cases, na.rm = TRUE), .groups = "drop")

weekly_trend |>
  pivot_wider(
    names_from = week,
    values_from = avg_new_cases,
    names_prefix = "week_of_"
  ) |>
  clean_names() |>
  print(max_footer_lines = 1)
```

## Putting it together

Each step above modified `covid_clean` and saved it back. Written as
one pipeline, the whole cleaning pass is:

```{r}
#| eval: false

covid_clean <-
  covid_data |>
  clean_names() |>
  mutate(
    last_update_date = mdy(last_update_date),
    town = town |> str_trim() |> str_to_title()
  ) |>
  arrange(town, last_update_date) |>
  group_by(town) |>
  mutate(new_cases = total_cases - lag(total_cases)) |>
  ungroup()
```

## Saving our data artifact

```{r}
#| eval: false
write_csv(covid_clean, here("data", "processed", "covid_clean.csv"))
```

Every future chapter can now start from `covid_clean` instead of
re-deriving it — one clean pass, reused everywhere.

::: {.callout-note}
Want to compare against a reference? See
[`project-snapshots/after-ch-02`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/after-ch-02)
on GitHub for the `scripts/` this chapter should leave you with.
:::
