---
title: "Describing and Testing"
packages: [dplyr, janitor]
---

```{r}
#| label: chapter-root
#| include: false
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

```{r}
#| label: load-linelist
#| include: false
linelist <- read_book_data("synthetic-covid-linelist.csv")
```

You built a line-list in the last chapter — one row per case, with an age, a
set of dates, and outcome flags. Now we get to *ask questions of it*. This
chapter is the everyday inferential toolkit: describe what's in the data,
cross-tabulate one variable against another, and run the handful of statistical
tests that answer "is this pattern real, or could it be chance?"

If you're coming from SAS, this is `PROC FREQ` and `PROC MEANS` territory —
just tidier, and with the whole thing scriptable end to end.

::: {.callout-warning}
## This analysis runs on fake data

Everything below runs on the **synthetic** line-list from the
[previous chapter](../building-a-dataset/index.qmd). The town names and case
counts are real, but every age, sex, date, and outcome was randomly generated.
The *shape* of the patterns is realistic — deaths really do skew older — but
**no row is a real person.** Never report these numbers as real, and never let
one of these tables walk out the door as if it described actual patients.
:::

## Getting set up

We'll lean on two packages: `dplyr` for the summarising and `janitor` for the
epi-friendly tables. In your own `covid-briefing/` project you'd start by
reading the file you built:

```{r}
#| label: read-linelist-shown
#| eval: false
linelist <- read_csv(here("data", "synthetic-covid-linelist.csv"))
```

Then we recreate the `age_group` column from the previous chapter — most of the
questions in this chapter are about how outcomes vary *across ages*, and a
handful of ordered bands is far easier to tabulate than raw years:

```{r}
#| label: prep
library(dplyr)
library(janitor)

linelist <- linelist |>
  mutate(
    age_group = case_when(
      age < 18 ~ "0-17",
      age < 50 ~ "18-49",
      age < 65 ~ "50-64",
      age < 80 ~ "65-79",
      TRUE     ~ "80+"
    )
  )
```

## Section 1 — Describing what you have

Before any test, look at the data. The first question about almost any outcome
is simply *how often did it happen?* — counts and percents.

`janitor::tabyl()` is the tool for that. Hand it a column and it returns a tidy
table of counts and proportions; `adorn_pct_formatting()` turns the raw
proportions into readable percentages. If you've reached for `PROC FREQ` in
SAS, this is the same job.

```{r}
#| label: tabyl-died
linelist |>
  tabyl(died) |>
  adorn_pct_formatting()
```

That's your case fatality snapshot: about 2.6% of cases in this line-list died.
`tabyl()` is a data frame like any other, so it drops straight into a report or
a `gt` table later without any wrestling.

Counts answer "how many." For a continuous variable like age you want the
*shape* — where the middle sits and how far it spreads. `summarise()` gives you
whichever statistics you name, and the classic epi move is to compute them
**split by outcome** so you can compare. Here's "does age differ between people
who died and people who survived?":

```{r}
#| label: age-by-outcome
linelist |>
  group_by(died) |>
  summarise(
    n = n(),
    mean_age   = round(mean(age), 1),
    median_age = median(age),
    min_age    = min(age),
    max_age    = max(age)
  )
```

Read across the two rows and the story jumps out. Among the people who died,
the mean age is about 74.8 and the median 79; among survivors the mean is 41.4
and the median 41. Deaths skew old — decades older, on average. The `mean` and
`median` sitting close together within each group also tells you neither is
being dragged around by a few extreme outliers.

This one table is often the whole descriptive section of a briefing. But it
also raises the obvious follow-up: we can *see* that deaths are older, but is
the association between age and death strong enough to trust? That's what the
next section is for.

## Section 2 — Cross-tabulation with a statistical test

The natural next step is a **two-way table**: age group down the side, outcome
across the top. `tabyl()` takes two variables and builds exactly that, and the
`adorn_*` helpers layer on the pieces you want to see.

```{r}
#| label: crosstab
linelist |>
  tabyl(age_group, died) |>
  adorn_totals("row") |>
  adorn_percentages("row") |>
  adorn_pct_formatting() |>
  adorn_ns()
```

There's a lot packed into that pipe, so here's what each adornment does:

- `adorn_totals("row")` adds a totals row across the bottom.
- `adorn_percentages("row")` converts each cell to a **row** percentage — so
  each age group's numbers add to 100% *across* its own row.
- `adorn_pct_formatting()` formats those as percents.
- `adorn_ns()` puts the raw counts back in, in parentheses, so you never lose
  sight of how many people each percentage rests on.

Why row percents? Because the row is the age group, so the percentage in the
`Yes` column is the **case fatality rate within that age group** — exactly the
number an epidemiologist wants. Reading down that `Yes` column, the CFR climbs
steadily: roughly 0.1% for 0–17, 0.5% for 18–49, 1.2% for 50–64, 7.2% for
65–79, and 29.3% for the 80+ group. Nearly three in ten of the oldest cases
died. That's the association from Section 1, now resolved group by group.

Seeing a trend in a table and *establishing* it statistically are two different
things. A **chi-square test** asks whether the association between age group and
outcome could plausibly have arisen by chance if the two were actually
unrelated.

```{r}
#| label: chisq
age_died <- table(linelist$age_group, linelist$died)
chisq.test(age_died)
```

`table()` builds the plain contingency table (chi-square wants counts, not
percentages), and `chisq.test()` does the rest. The null hypothesis is that
there is **no association** — that outcome doesn't depend on age group. The test
returns a huge X-squared statistic (around 5176) on 4 degrees of freedom and a
p-value below 2.2e-16, which is R's way of printing "smaller than I bother to
show." In plain terms: if age and death were truly unrelated, a table this
lopsided would essentially never happen. We reject the null.

::: {.callout-warning}
## A tiny p-value is not a big effect

With this many cases, almost *any* association will come back "significant" —
the p-value is partly a measure of your sample size. A huge `n` makes the test
exquisitely sensitive, so it will flag even trivial differences. Don't let a
row of zeros after the decimal point fool you into thinking you've found
something important. The **effect size** — the actual jump from 0.1% to 29.3%
CFR you read off the cross-tab — is what makes this finding matter. The
p-value only tells you it's not noise; the percentages tell you it's a big
deal.
:::

### When cells are small: Fisher's exact test

Chi-square relies on an *approximation* that gets shaky when the expected count
in any cell is small — the usual rule of thumb is that no expected cell should
drop below 5. That happens all the time when you slice down to a single small
town or a rare outcome. **Fisher's exact test** computes the exact probability
instead of approximating it, so it stays honest on sparse tables.

Let's force the situation by looking at one small town, Hartland:

```{r}
#| label: fisher
hartland <- linelist |> filter(town == "Hartland")

table(hartland$hospitalized, hartland$died)

fisher.test(table(hartland$hospitalized, hartland$died))
```

Hartland has only 59 cases, and the `table()` shows why chi-square would
struggle: the cells are thin and at least one is a flat zero. `fisher.test()`
handles that without complaint and returns a p-value of about 0.0018 — strong
evidence of an association between being hospitalized and dying, even in this
tiny slice.

The teaching point: **reach for Fisher when the counts are small.** If you run
`chisq.test()` on a sparse table it'll warn you that the approximation may be
unreliable — that warning is your cue to switch. On small or zero-heavy tables,
Fisher is the trustworthy choice.

## Section 3 — Testing a trend in proportions

The stakeholder didn't just ask "is there an association" — they asked about
**trends in proportions**, like an outcome rising over time or across ordered
categories. A plain chi-square treats the age groups as unordered labels; it
would give the same answer if you shuffled them. But our age bands have a
natural order, and the interesting claim is specifically that CFR *rises
monotonically* from young to old.

The **test for trend in proportions** (a Cochran–Armitage-style test) is built
for that. It needs two things in order: the number of deaths in each group, and
the number of cases in each group. We build those with a grouped
`summarise()`, then `arrange()` to lock in the order:

```{r}
#| label: trend
cfr_by_age <-
  linelist |>
  group_by(age_group) |>
  summarise(deaths = sum(died == "Yes"), n = n()) |>
  arrange(age_group)

cfr_by_age

prop.trend.test(cfr_by_age$deaths, cfr_by_age$n)
```

`sum(died == "Yes")` counts the deaths in each group (a `TRUE`/`FALSE` vector
sums to the number of `TRUE`s), and `n()` counts everyone. `prop.trend.test()`
takes the deaths and the group sizes and asks: is there a **directional**
trend in the proportion across these ordered groups?

The result — X-squared around 2919 on 1 degree of freedom, p below 2.2e-16 —
says yes, emphatically: CFR increases in a clear linear fashion across the age
bands. Because this test uses the *ordering* of the groups, it's a sharper
instrument than a plain chi-square when the order is meaningful. It's asking a
more specific question ("does it trend?") and so gives stronger evidence when
the answer is yes.

The same machinery answers the stakeholder's "outcome by year" question
directly: group by year instead of age band, feed the yearly death counts and
case counts into `prop.trend.test()` **in chronological order**, and you're
testing whether the outcome trended up or down over time. Any ordered grouping
— age bands, years, weeks of a wave — works the same way.

## Wrap-up

You can now do the three things that make up the bread-and-butter of case-level
analysis: **describe** a line-list with counts and summaries, **cross-tabulate**
two variables and test the association with chi-square or Fisher's exact test,
and **test a trend** across ordered groups. That's the everyday inferential
toolkit, and it'll carry you through most routine surveillance questions.

What it *won't* do is untangle variables that move together. Older cases might
also be more likely to be hospitalized, or to belong to a particular group — and
these simple tests look at one relationship at a time. The next chapter,
**Modeling**, is how you get *adjusted* effect estimates that hold age, sex, and
the rest constant all at once.

::: {.callout-note}
Want this chapter's work as a runnable script? See
[`project-snapshots/line-list-practice`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/line-list-practice)
on GitHub. `02_describe_and_test.R` is the script version of everything above —
the tabyls, the chi-square, Fisher's, and the trend test — ready to drop into
your `scripts/` folder.
:::
