---
title: "Modeling: Adjusted Effect Estimates"
packages: [dplyr, broom]
---

```{r}
#| label: chapter-root
#| include: false
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

```{r}
#| label: load-linelist
#| include: false
linelist <- read_book_data("synthetic-covid-linelist.csv")
```

The [previous chapter](describing-and-testing.qmd) left you with a cliffhanger.
Your cross-tabs showed two clear things: case fatality climbs steeply with age,
and — separately — hospitalized cases die far more often than non-hospitalized
ones. Both are real signals. But they're tangled together. Older people are
*also* the ones most likely to be hospitalized, so when you look at
hospitalization and death in a cross-tab, some of what you're seeing is really
just age wearing a different hat.

A two-way table can't pull those apart. It only ever looks at one relationship
at a time. What you need is a way to ask: *what is the effect of age on death,
holding hospitalization and sex constant?* And at the same time, *what is the
effect of hospitalization, holding age and sex constant?* That "holding the
others constant" is exactly what a **regression** does, and it's exactly what a
stakeholder means when they ask for **adjusted effect estimates**.

This chapter fits a **logistic regression** — the workhorse model for a
yes/no outcome like *died* — and reads the results as **odds ratios** with
confidence intervals. If this is your first model in R, don't worry: we'll take
it a piece at a time and translate every number into a plain sentence.

::: {.callout-warning}
## This models fake data

Everything here runs on the **synthetic** line-list from the
[Building a Dataset](../building-a-dataset/index.qmd) chapter. The town names
and case counts are real, but every age, sex, hospitalization, and death was
randomly generated. The patterns are real-*shaped* — deaths really do skew old
— but **no row is a real person.** Never report one of these odds ratios as a
real finding. They exist only so you can practice the mechanics on data that
behaves like the real thing.
:::

## From cross-tab to model

The previous chapter's tools — chi-square, Fisher's, the trend test — each
answer a question about *one* pair of variables. Regression answers a different
kind of question: given *several* predictors at once, what does each one
contribute on its own? The model estimates every effect **net of the others**,
which is the whole point. When it tells you the odds ratio for age, that number
already has hospitalization and sex accounted for.

::: {.callout-note}
## This chapter uses broom

We'll use **broom** to turn the model's output into a tidy data frame. It may
not be installed yet. In your own `covid-briefing/` project, install it once
with:

```r
pak::pak("broom")
```

If you're building this book from its repository (which uses `renv`), install
it into the project library instead and record it in the lockfile:

```r
renv::install("broom")
renv::snapshot()
```

`broom` takes the somewhat awkward object that `glm()` hands back and turns it
into a clean, rectangular data frame you can read, filter, and drop into a
report — the same shape as everything else you've worked with.
:::

## Fit a logistic regression

A little prep first. Two small but important setup moves before we fit
anything.

**First, an ordered age variable with a deliberate reference level.** A
regression with a categorical predictor doesn't estimate an effect for *every*
level — it picks one level as the baseline (the "reference") and reports every
other level *relative to it*. We make `age_group` a `factor` and put `"18-49"`
first, which makes it the reference. That's a choice with meaning: healthy
working-age adults are the natural comparison group, so every age odds ratio
will read as "compared to a 18–49-year-old." Reference-first is a habit worth
building.

**Second, a numeric 0/1 outcome.** Logistic regression models the probability
of an *event* happening, so it needs the outcome coded as `1` for the event
(died) and `0` for not. Our `died` column is `"Yes"`/`"No"` text;
`as.integer(died == "Yes")` turns it into exactly the `1`/`0` the model wants.

```{r}
#| label: prep
library(dplyr)
library(broom)

model_data <-
  linelist |>
  mutate(
    age_group = factor(
      case_when(
        age < 18 ~ "0-17",
        age < 50 ~ "18-49",
        age < 65 ~ "50-64",
        age < 80 ~ "65-79",
        TRUE     ~ "80+"
      ),
      levels = c("18-49", "0-17", "50-64", "65-79", "80+")
    ),
    died_flag = as.integer(died == "Yes")
  )
```

Now the model itself. `glm()` — **g**eneralized **l**inear **m**odel — is the
function, and the two arguments that make it a *logistic* regression are the
formula and the `family`.

The formula `died_flag ~ age_group + sex + hospitalized` reads left to right as
"**died explained by** age group, plus sex, plus hospitalization." The `~` is
"is modeled by"; the `+` signs add predictors. `family = binomial` is the piece
that makes this logistic: it tells `glm()` the outcome is a yes/no event and to
work on the **log-odds** scale rather than fitting a straight line.

```{r}
#| label: fit
model <- glm(
  died_flag ~ age_group + sex + hospitalized,
  data = model_data,
  family = binomial
)

summary(model)
```

`summary()` prints the full model report — the same one SAS's `PROC LOGISTIC`
would give you, if you're coming from that world. There's a lot there, but the
key thing to know right now is that the numbers in the `Estimate` column are on
the **log-odds** scale. A log-odds coefficient is nearly impossible to read
directly — nobody thinks in log-odds. That's why we almost never stop here.
The next step converts these into **odds ratios**, which you *can* read.

## Odds ratios with broom

Here's where `broom` earns its keep. `tidy()` takes the model and returns a
tidy data frame, one row per term. Two arguments do the heavy lifting:
`exponentiate = TRUE` converts those log-odds coefficients into **odds ratios**,
and `conf.int = TRUE` adds a **95% confidence interval** for each one.

```{r}
#| label: odds-ratios
tidy(model, exponentiate = TRUE, conf.int = TRUE)
```

Now every number is legible. The `estimate` column is the odds ratio, and
`conf.low`/`conf.high` bracket the 95% CI. Here's how to read an odds ratio:

- **Greater than 1** means *higher* odds of the outcome than the reference.
- **Less than 1** means *lower* odds.
- **Exactly 1** means no difference at all.

And crucially — every one of these is **adjusted** for the other predictors in
the model. Walk the age gradient, each compared to the 18–49 reference group:

- **`age_group0-17`: 0.19** (95% CI 0.06–0.45). Children have about a fifth the
  odds of death of a working-age adult — far *lower*, which is why the OR sits
  well below 1.
- **`age_group50-64`: 2.19** (1.62–2.95). Roughly double the odds.
- **`age_group65-79`: 12.1** (9.44–15.6). Twelve times the odds.
- **`age_group80+`: 55.1** (43.2–71.1). Fifty-five times the odds of death of a
  working-age adult — a dramatic age gradient, and all of it *after* adjusting
  for whether the person was hospitalized.

Then the other two predictors:

- **`hospitalizedYes`: 4.10** (3.49–4.82). Hospitalized cases have about four
  times the odds of death — and, importantly, this is the hospitalization
  effect *with age already accounted for*. It's not just older people showing
  up again in disguise.
- **`sexMale`: 1.03** (0.89–1.19). This one is the teaching moment. The odds
  ratio is a hair above 1, but look at the confidence interval: it runs from
  0.89 to **1.19** — it *crosses 1*.

That last point deserves its own paragraph. When a confidence interval crosses
1, the data can't distinguish the effect from "no effect at all." The interval
includes odds ratios below 1 (protective) *and* above 1 (harmful), so you
genuinely cannot say which direction, if any, sex pushes the outcome. This is a
**null adjusted estimate**, and it's a clean, concrete example of one. Contrast
it with the previous chapter's habit of eyeballing a cross-tab and declaring a
pattern: here the model gives you a principled way to say "no, there's nothing
to see in this variable" — the interval says so.

A nice sanity check falls out of all this. Remember *how* the synthetic data
was built: with a strong age effect baked in, no sex effect, and a
hospitalization bump. The model **recovered exactly that structure** — a steep
age gradient, a null result for sex, a moderate hospitalization effect. When a
model reproduces the signal you know is in the data, that's reassurance about
both the model *and* the data build.

::: {.callout-warning}
## An odds ratio is not a risk ratio

People often read an odds ratio as if it were a *risk* ratio ("55× the risk"),
but the two only agree when the outcome is **rare**. For a common outcome the
odds ratio exaggerates the risk ratio, sometimes badly. Death is uncommon
enough here that the gap is modest, but keep the distinction in mind — say
"odds," not "risk," when you report an OR.
:::

## The tidymodels way

The `glm()` you just ran is the right tool for this job. When your goal is an
**explanatory** model — you want an interpretable effect estimate, an adjusted
odds ratio to hand a stakeholder — plain `glm()` plus `broom` is lean, clear,
and hard to beat.

There's another whole ecosystem in the tidyverse for modeling, called
**tidymodels**, and it's worth knowing where it fits. Its real strength is
**prediction**: when the goal isn't "explain" but "forecast," tidymodels gives
you train/test splits, resampling, hyperparameter tuning, and standardized
performance metrics, all with a consistent interface. That's a different job
from what we did above.

But two of tidymodels' habits are worth borrowing even for an explanatory
model: **recipes** for declaring your preprocessing once, and **parsnip** for a
model spec you can re-point at a different engine without rewriting. Here's the
same logistic regression, expressed the tidymodels way. We're showing it as a
reference — it does **not** run when the book builds — so you can read the shape
of it without needing tidymodels installed:

```{r}
#| label: tidymodels-way
#| eval: false
library(recipes)
library(parsnip)
library(workflows)

# died as a factor outcome for classification; first level is the reference
model_data <- model_data |>
  mutate(died = factor(died, levels = c("No", "Yes")))

# 1. recipe: declare outcome + predictors and preprocessing (dummy-code factors)
death_recipe <-
  recipe(died ~ age_group + sex + hospitalized, data = model_data) |>
  step_dummy(all_nominal_predictors())

# 2. parsnip spec: logistic regression via the glm engine
death_spec <-
  logistic_reg() |>
  set_engine("glm")

# 3. workflow bundles preprocessing + model and fits in one place
death_fit <-
  workflow() |>
  add_recipe(death_recipe) |>
  add_model(death_spec) |>
  fit(data = model_data)

# same tidy odds-ratio table, now from the workflow
tidy(death_fit, exponentiate = TRUE, conf.int = TRUE)
```

What each piece buys you:

- **recipes** — reproducible preprocessing, declared once and reused. Here it's
  just dummy-coding the factors, but the same idea scales: `step_other()` to
  collapse rare categories into an "other" bucket, `step_normalize()` to center
  and scale continuous predictors, and dozens more. You write the recipe once
  and every fit and prediction uses the identical steps.
- **parsnip** — one model spec, swappable engines. `logistic_reg()` is the
  *kind* of model; `set_engine("glm")` picks the implementation. Want a
  penalized version instead? Swap in `set_engine("glmnet")` and you don't
  rewrite the fit.
- **workflows** — bundle the recipe and the model together so they can't drift
  apart. The preprocessing and the model travel as one object, which is a
  genuine reproducibility win on a bigger project.

**When should you actually reach for full tidymodels?** When the goal is
prediction. That's when you'll add `rsample` for train/test splits and
cross-validation, and `yardstick` for metrics like ROC-AUC and accuracy. For an
adjusted odds ratio, though — a single explanatory model you want to interpret —
`glm()` and `broom` are lighter and clearer, and there's no prize for
over-tooling. The full prediction track lives at
<https://www.tidymodels.org/>.

One gotcha to file away: `logistic_reg()` needs a **factor** outcome (not the
0/1 integer `glm()` wanted), and tidymodels treats the **first** factor level as
the reference — the *non*-event. That's why the code above sets the levels as
`c("No", "Yes")` deliberately, so that `"Yes"` (death) is the event being
modeled. Set your factor levels on purpose; don't let the default alphabetical
order decide which outcome you're predicting.

## Wrap-up

You fit a real adjusted model. You took a tangle that a cross-tab couldn't
resolve — is it age, or is it hospitalization? — and got a clean answer for
each: an odds ratio, adjusted for the others, with a confidence interval that
tells you how much to trust it. You saw a strong effect (age), a moderate one
(hospitalization), and a genuinely null one (sex, its CI straddling 1), and you
learned to read all three off a single tidy table. And you saw where the
tidyverse's bigger modeling toolkit fits, and where it's more machine than the
job needs.

One honest gap remains. Everything here is about **odds** — the odds that a
*known case* died. It says nothing about **rates** in a *population*:
age-adjusted death rates, incidence-rate ratios, the per-100,000 measures that
much of surveillance runs on. Those need a **population denominator** — how many
people live in each town and age band — which this book doesn't have in its data
yet. When a denominator dataset lands, a future chapter can take up rate-based
measures. Until then, you've got the case-level model that answers the
stakeholder's actual question: the adjusted effect estimates.

::: {.callout-note}
Want this chapter as a runnable script? See
[`project-snapshots/line-list-practice`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/line-list-practice)
on GitHub. `03_model_deaths.R` is the script version of the logistic regression
and odds-ratio table you just built, ready to drop into your `scripts/` folder.
:::
