---
title: Managing Secrets
packages: [dotenv, keyring, httr2]
---

```{r}
#| label: chapter-root
#| include: false
source(here::here("preamble-packages.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

## What are Secrets {#sec-secrets}

Secrets are pieces of information - API keys, passwords, tokens, connection
strings - that grant access to something and should never end up in your
script, your git history, or your rendered output. The moment a secret is
typed directly into a `.R` or `.qmd` file, it is one `git push` away from
being public forever, even if you delete it in a later commit. History
remembers.

The fix isn't complicated: keep secrets *out* of code and load them in at
runtime from somewhere code doesn't normally look - an environment
variable.

## `.Renviron`

R users will often bump into `.Renviron` first. It's a plain text file,
usually living in your project root or your home directory, that R reads
automatically on startup and turns into environment variables.

```{bash}
#| eval: false

# .Renviron
CENSUS_API_KEY=abc123yourkeyhere
```

Restart your R session and the value is available via `Sys.getenv()`:

```{r}
#| eval: false

Sys.getenv("CENSUS_API_KEY")
```

`.Renviron` works, but it's R-specific - the skill doesn't transfer. Learn
this pattern and it's only useful the next time you open RStudio or
Positron. For that reason we're going to spend most of this chapter on
`.env` instead, a convention almost every language and tool understands.

## `.env` and the `dotenv` package

A `.env` file is the same idea as `.Renviron` - `KEY=value` pairs, one per
line - but it isn't R-specific. Python, Node, Docker, and most CI systems
all read `.env` files the same way. Learn it once here, reuse it
everywhere else in your career.

```{bash}
#| eval: false

# .env
CENSUS_API_KEY=abc123yourkeyhere
```

::: {.callout-important}
Add `.env` to your `.gitignore` before you ever put a real value in it.
An untracked file can't leak.
:::

R doesn't read `.env` automatically the way it reads `.Renviron`, so we
install a small package to bridge the gap.

```{r}
#| eval: false

install.packages("dotenv")
```

`dotenv::load_dot_env()` reads the file and populates the environment
for the current session:

```{r}
#| eval: false

library(dotenv)
load_dot_env(".env")

Sys.getenv("CENSUS_API_KEY")
```

### Worked example: the Census API key

The [US Census Bureau's API](https://api.census.gov/data/key_signup.html)
is a good real-world case - it's free, it's public-health relevant, and it
requires a key. Sign up, drop the key in `.env` as `CENSUS_API_KEY`, and
never type it into a script again.

```{r}
#| eval: false

library(dotenv)
library(httr2)

load_dot_env(".env")

census_key <- Sys.getenv("CENSUS_API_KEY")

request("https://api.census.gov/data/2020/dec/pl") |>
  req_url_query(
    get = "P1_001N,NAME",
    `for` = "state:09", # Connecticut
    key = census_key
  ) |>
  req_perform() |>
  resp_body_json()
```

Anyone reading this script sees that a key is required and where it comes
from - but never the key itself. Share the script freely; the `.env` file
stays on your machine.

## `keyring`: secrets outside any file

`.env` still means a secret sits somewhere on disk in plain text. The
[`keyring`](https://keyring.r-lib.org/) package goes a step further by
storing secrets in your operating system's own credential store - Keychain
on macOS, Credential Manager on Windows, the Secret Service on Linux.
Nothing is ever written to a project file at all.

```{r}
#| eval: false

install.packages("keyring")
library(keyring)

# store it once, interactively - you'll be prompted, it won't appear
# in your script or console history
key_set("census_api_key")

# retrieve it whenever you need it
census_key <- key_get("census_api_key")
```

`keyring` is a good fit for secrets you use across many projects on your
own machine - a personal API key, a database password - where per-project
`.env` files would mean re-entering the same value over and over.

## `httr2::secret_encrypt()`: sharing secrets safely in code

`.env` and `keyring` both assume the secret lives *outside* what you
share. Sometimes that's not enough - a package or a shared script needs
to ship with a credential built in, for a demo API or a service account
that isn't sensitive on its own but shouldn't be plain text either.
`httr2` provides `secret_encrypt()` for exactly this case.

The idea: encrypt the secret with a key, commit the *encrypted* value,
and keep the encryption key itself in an environment variable that never
gets committed.

```{r}
#| eval: false

library(httr2)

# generate an encryption key once, store it as an env var
# (e.g. MY_PACKAGE_KEY in .Renviron or .env - never in code)
secret_make_key()
```

```{r}
#| eval: false

# encrypt the real secret using that key
encrypted_census_key <- secret_encrypt(
  "abc123yourkeyhere",
  "MY_PACKAGE_KEY"
)

encrypted_census_key
#> [1] "encrypted:5f8a2c..."
```

That `encrypted:...` string is safe to commit - it's useless without the
`MY_PACKAGE_KEY` environment variable to decrypt it. At request time,
decrypt on the fly:

```{r}
#| eval: false

request("https://api.census.gov/data/2020/dec/pl") |>
  req_url_query(key = secret_decrypt(encrypted_census_key, "MY_PACKAGE_KEY"))
```

This is the pattern `httr2` itself recommends for package authors who
need to ship example code that actually runs in CI without exposing the
underlying credential. For a deeper dive, see the `httr2` article on
[wrapping APIs](https://httr2.r-lib.org/articles/wrapping-apis.html#secret-management).

## Picking an approach

- **One-off project, R only:** `.Renviron` is fine.
- **One-off project, skill you want to transfer beyond R:** `.env` +
  `dotenv`.
- **Secret reused across many projects on your machine:** `keyring`.
- **Secret that needs to travel with shared code or a package:**
  `httr2::secret_encrypt()`.

Whichever you choose, the rule doesn't change: secrets live in the
environment, not in the script.
