---
title: "Checkpoint: Where the Project Stands"
packages: [fs, glue]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
library(dplyr)
library(readr)
library(janitor)
library(lubridate)
library(glue)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

No new package today, no new geometry or verb to learn. This far in,
it's worth stopping to combine everything you've built so far into one
thing the Commissioner could actually read.

## The arc so far

| Part | Question it answered | What it left in `covid-briefing/` |
|---|---|---|
| 00 — Orientation | Why R, and how is the project folder organized? | The scaffold itself, and `data/raw/` with the CT COVID CSV |
| 01 — Working with Real Data | How do I read, inspect, and trust this file? | A Hartford-only tibble, filtered and column-renamed by hand |
| 02 — Cleaning Data | How do I fix names, dates, and messy text? | `covid_clean` — every column name fixed, dates parsed, town names trimmed and title-cased |
| 03 — Data Visualization | How do I show this to someone who won't open a CSV? | A trend chart and a statewide map, saved to `outputs/figures/` |

Each chapter's project picked up exactly where the last one left off.
That's on purpose — a real analysis rarely restarts from scratch every
time a new technique gets introduced.

## This week's briefing summary

Charts are great for a slide deck, but sometimes the Commissioner just
wants the numbers in a sentence, right now, before the meeting starts.
That's today's artifact: a short status paragraph built entirely from
skills you already have — `dplyr` to find the numbers, `lubridate` to
know which date counts as "now", and `glue` to turn it into English.

```{r}
#| label: status-setup
#| include: false
covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")

covid_clean <-
  covid_data |>
  clean_names() |>
  mutate(last_update_date = mdy(last_update_date))
```

```{r}
#| label: status-numbers
latest_date <- max(covid_clean$last_update_date)
days_of_data <- as.numeric(latest_date - min(covid_clean$last_update_date))

statewide_latest <-
  covid_clean |>
  filter(last_update_date == latest_date) |>
  summarise(
    total_cases = sum(total_cases, na.rm = TRUE),
    total_deaths = sum(total_deaths, na.rm = TRUE)
  )

hartford_latest <-
  covid_clean |>
  filter(town == "Hartford", last_update_date == latest_date) |>
  select(total_cases, total_deaths)
```

```{r}
#| label: status-summary
briefing_summary <- glue(
  "Status as of {latest_date} ({days_of_data} days of data collected):\n",
  "Statewide: {format(statewide_latest$total_cases, big.mark = ',')} total cases, ",
  "{format(statewide_latest$total_deaths, big.mark = ',')} total deaths.\n",
  "Hartford: {format(hartford_latest$total_cases, big.mark = ',')} total cases, ",
  "{format(hartford_latest$total_deaths, big.mark = ',')} total deaths."
)

briefing_summary
```

Four lines of code, no new concepts, and a paragraph that's actually
correct as of whatever the latest date in the file happens to be —
because it's computed, not typed in by hand.

### Saving the checkpoint artifact

```{r}
#| label: save-summary
#| eval: false
write_lines(
  briefing_summary,
  here("outputs", "tables", "briefing_summary.txt")
)
```

That text file is this checkpoint's deliverable, the same way a chart
was chapter 3's — a small, real thing that lives in the project folder
and didn't exist before today.

## What your project folder should look like now

```{r}
#| eval: false
fs::dir_tree(here::here(), recurse = 2)
```

If you've been following along in your own `covid-briefing/` project,
you should see something close to this:

```
covid-briefing/
├── data/
│   ├── raw/
│   │   └── ct_covid_by_town_2026-06-30.csv
│   └── processed/
│       └── covid_clean.csv        # if you uncommented the write_csv() in ch. 2
├── R/
├── scripts/
├── outputs/
│   ├── figures/
│   │   ├── hartford_new_cases.png
│   │   ├── hartford_vs_state_trend.png
│   │   └── statewide_case_rate_map.png
│   └── tables/
│       └── briefing_summary.txt
└── docs/
```

If your folder looks different, that's fine — the point isn't to match
this tree exactly, it's to notice that the folder is *filling up with
real artifacts* instead of staying empty. That's what progress looks
like in a data project: not lines of code, but files someone else could
open and use.

## What's still manual

Being honest about the current state of things:

- Every chapter's project re-runs the cleaning steps by hand, from the
  top, in a `.qmd` file. Nothing here is a function you could call
  twice.
- The raw CSV is a snapshot someone downloaded once. Nothing fetches
  new data automatically.
- "Hartford" and the current date are typed directly into the code.
  Asking about a different town means editing the script, not passing
  an argument.
- There's no history of any of this yet — if you overwrite a file by
  accident, it's just gone.

None of that is a problem yet — it's normal for a project this size.
It becomes a problem the day someone asks for the same briefing next
week, for a different town, without you sitting at the keyboard.

## What's next

The parts ahead exist to close exactly those gaps:

- **Managing Code** — version control, code style, and documentation,
  so this project can be shared, reviewed, and trusted by someone
  other than you.
- **Intermediate R: Automating the Manual Steps** — turning repeated
  code into functions, and pulling fresh data straight from the state's
  API instead of re-downloading a CSV by hand.

The scenario hasn't changed: the Commissioner still wants to know if
it's getting worse in Hartford. What changes from here is how much of
the answer you have to rebuild yourself every time someone asks.
