---
title: "Working with Excel Files"
---

Excel files are everywhere in public health. This chapter covers reading and
writing `.xlsx` files using `readxl` and `writexl`.

## Installing the packages

```{r}
#| eval: false

install.packages(c(
  "readxl",  # for reading .xlsx files
  "writexl"  # for writing .xlsx files
))
```

## Reading `.xlsx` files

`readxl` reads Excel files the same way `readr` reads CSVs. Use `here()` to
build the path so it works both interactively and during render.

```{r}
#| eval: false

library(readxl)
library(here)

xlsx_path <- here("data", "raw", "your-file.xlsx")
your_data <- read_excel(xlsx_path)

your_data
```

If your workbook has multiple sheets, use the `sheet` argument:

```{r}
#| eval: false

your_data <- read_excel(xlsx_path, sheet = "Sheet2")

# or by position
your_data <- read_excel(xlsx_path, sheet = 2)
```

## Writing `.xlsx` files

`writexl` writes a data frame to an Excel file with no Java dependency
(unlike `openxlsx`).

```{r}
#| eval: false

library(writexl)
library(here)

output_path <- here("data", "my-output.xlsx")

write_xlsx(your_data, path = output_path)
```

To write multiple sheets in one workbook, pass a named list:

```{r}
#| eval: false

write_xlsx(
  list(
    cases  = cases_data,
    deaths = deaths_data
  ),
  path = output_path
)
```

## When to use CSV vs Excel

Prefer CSV for anything that feeds into analysis — it is plain text, version-control
friendly, and readable by every tool. Use Excel output when the end recipient
is a stakeholder who will open the file directly in Excel and expects formatting.
