---
title: "Joins"
---

The choropleth section used `left_join()` to attach COVID case data to
town boundaries. That's one of six join verbs `dplyr` gives you — each
answers a different question about how two tables relate. This page
walks through all six using the same shape of data: a `towns` table
(the "left" table, like `ct_towns`) and a `cases` table (the "right"
table, like `town_cases`).

```{r}
#| label: joins-setup
library(dplyr)

towns <- tibble::tibble(
  name = c("Hartford", "New Haven", "Bridgeport", "Norwalk"),
  county = c("Hartford", "New Haven", "Fairfield", "Fairfield")
)

cases <- tibble::tibble(
  name = c("Hartford", "New Haven", "Stamford"),
  case_rate = c(412, 388, 355)
)
```

`towns` has four rows, `cases` has three, and only two names
(`Hartford`, `New Haven`) appear in both. `Norwalk` has no case data.
`Stamford` has case data but isn't in `towns`. That mismatch is exactly
what makes joins interesting — each verb handles it differently.

## `left_join()`: keep everything on the left

```{r}
#| label: left-join
towns |>
  left_join(cases, by = "name")
```

Every row from `towns` survives. Where `cases` has no matching `name`
(Norwalk), `case_rate` comes back `NA`. `Stamford` never appears — it
wasn't in the left table to begin with. This is what the choropleth
uses: keep all 169 CT towns, fill in case data where it exists, show
gray for the rest.

## `right_join()`: keep everything on the right

```{r}
#| label: right-join
towns |>
  right_join(cases, by = "name")
```

The mirror image. Every row from `cases` survives, including
`Stamford`, which now shows `NA` for `county`. Norwalk drops out
entirely since it's not in `cases`. In practice, `right_join(x, y)` is
the same as `left_join(y, x)` with the columns reordered — most people
just pick whichever table they name first and use `left_join()`.

## `inner_join()`: keep only matches

```{r}
#| label: inner-join
towns |>
  inner_join(cases, by = "name")
```

Only rows present in **both** tables survive: Hartford and New Haven.
No `NA`s, because everything left in the result has a match on both
sides. Use this when a row without a match is meaningless for what
you're calculating — e.g., a rate that requires both a population and
a case count.

## `full_join()`: keep everything from both

```{r}
#| label: full-join
towns |>
  full_join(cases, by = "name")
```

Nothing is dropped. Norwalk keeps its row with `case_rate` as `NA`;
Stamford keeps its row with `county` as `NA`. Use this when you want a
complete picture of every name that shows up anywhere, gaps and all.

## `semi_join()`: filter, don't add columns

```{r}
#| label: semi-join
towns |>
  semi_join(cases, by = "name")
```

`semi_join()` keeps rows from `towns` that **have** a match in `cases`
— but it never adds `cases`' columns. The result has only `name` and
`county`. Think of it as a filter, not a join: "give me the towns that
also appear in my case data."

## `anti_join()`: the opposite filter

```{r}
#| label: anti-join
towns |>
  anti_join(cases, by = "name")
```

The inverse of `semi_join()`: keep rows from `towns` that have **no**
match in `cases`. This is the fastest way to answer "which towns are
missing case data?" — exactly the gray towns in the choropleth.

::: {.callout-tip}
## Picking the right join

- Need every row from your main table, filled in where possible? →
  `left_join()`
- Need only rows that exist on both sides? → `inner_join()`
- Need every row from either side, gaps and all? → `full_join()`
- Just need to filter one table by presence in another, without
  adding columns? → `semi_join()` / `anti_join()`
:::
