---
title: "Reading Data: Formats and Gotchas"
---

Data shows up in a lot of shapes. A CSV emailed from a hospital, an Excel
workbook from a lab, a Microsoft Access database that's been passed around
a program for a decade. R can read all of these, but a stakeholder's very
fair question is: *once it's in R, does the data need reformatting, and does
that depend on the file type?*

The short version: the file *type* changes **how** you read it — which package
and function you reach for — but most of the cleanup headaches are the **same**
no matter where the data came from. Dates come in as text, column names arrive
messy, and blank cells mean different things to different systems. Those quirks
follow the data around regardless of format.

This page is the reference for those cross-cutting quirks. For the actual
*how-to* of reading files, you've already seen it:

- [Loading Your First Dataset](../01-working-with-real-data/index.qmd) covers
  reading a `.csv` with `readr::read_csv()`.
- [Working with Excel Files](excel-files.qmd) covers `.xlsx` with
  `readxl::read_excel()`.

This page is about the *gotchas* that come after the read.

## Formats at a glance

Here are the formats an epidemiologist meets most often, the function that reads
each one, and the quirks that come with the territory.

| Format | Read it with | Format-specific quirks |
|---|---|---|
| `.csv` (comma-separated) | `readr::read_csv()` | Plain text, so there's no stored notion of "this is a date" or "this is a number" — `readr` guesses column types from the values. Watch for commas *inside* fields, and for regional files that use `;` as the separator (`read_csv2()`). |
| `.xlsx` / `.xls` (Excel) | `readxl::read_excel()` | Multiple sheets (use `sheet =`), merged cells, and title rows above the real header. Excel stores dates as **serial numbers** under the hood, which sometimes surface as numbers like `44927` instead of a date. |
| `.accdb` / `.mdb` (MS Access) / other databases | `DBI` + `odbc`, **or** export to CSV first | Needs a database connector and a platform-specific driver — often Windows-only. See the note below. |

Databases deserve an honest note rather than a full tutorial. Reading directly
from Microsoft Access (or SQL Server, or similar) means connecting to it with
the `DBI` package plus a backend like `odbc`, and that in turn needs a
database *driver* installed on your machine — which is frequently
Windows-specific and not always something you can install yourself. The
pragmatic path, and the one worth trying first, is to open the database and
**export the table you need to CSV**, then read that CSV with the tools you
already know. A fuller treatment of connecting to databases from R is a good
candidate for a later chapter.

## The gotchas

### Dates read in as text

This is the one that trips up almost everyone. Unless you tell it otherwise, R
will read a column of dates as plain **character** text. A value like
`06/30/2026` is just a string of characters as far as R is concerned — you
can't do date math on it, sort it chronologically, or filter a date range until
it's been *parsed* into a real date.

Parsing is exactly what [`lubridate`](../02-cleaning-data/index.qmd) is for. You
tell it the order the pieces appear in — month, day, year — and it handles the
rest:

```{r}
#| eval: false
library(lubridate)

mdy("06/30/2026") # month-day-year
ymd("2026-06-30") # year-month-day
```

There's no format string to memorize; the function name spells out the order.
The [Cleaning Data chapter](../02-cleaning-data/index.qmd) walks through this on
the real COVID dataset.

::: {.callout-warning}
## Excel's date serial-number trap
Excel doesn't store dates as text at all — it stores them as a **serial
number** counting days from an internal start date. Most of the time `readxl`
converts these back to dates for you, but if a column has mixed content, or the
cell was formatted oddly, dates can arrive in R as bare numbers like `44927`.

If you open a "date" column and find integers around `40000`–`46000`, that's
almost certainly an Excel serial number. You can convert it back with
`as.Date(44927, origin = "1899-12-30")`. When you spot this, it's worth going
back to the source workbook to confirm what those numbers are supposed to mean
before trusting the conversion.
:::

### Variable (column) names

R is picky about names, and data files rarely respect its rules. A "clean" R
name — what R calls a *syntactic* name — must:

- **not start with a number** (`2020_cases` is illegal; `cases_2020` is fine),
- **contain no spaces** (`Total Cases` breaks),
- **avoid most punctuation** (`%`, `#`, `/`, `(` and friends are out;
  underscores and dots are allowed),
- and remember that names are **case-sensitive** (`Town` and `town` are two
  different things).

R won't refuse to *hold* a messy name — it just forces you to wrap it in
backticks every single time you refer to it:

```{r}
#| eval: false
# a column literally named "Total Cases" with a space
covid_data |>
  select(`Total Cases`)
```

Typing backticks around every reference gets old fast. The fix is
`janitor::clean_names()`, taught in the
[Cleaning Data chapter](../02-cleaning-data/index.qmd). It lowercases
everything, swaps spaces for underscores, and strips the problem characters in
one call:

```{r}
#| eval: false
library(janitor)

covid_data <- clean_names(covid_data)
# "Total Cases" becomes total_cases, no backticks needed ever again
```

On length: R technically allows very long names, but keep them short and
readable anyway. Some export formats and downstream tools quietly truncate long
names, and a column called `n_confirmed` is easier to live with than
`number_of_confirmed_cases_reported_this_week`.

### Case sensitivity bites twice

Case sensitivity isn't just a naming rule — it applies to the **data values**
too. To R, `"Hartford"`, `"hartford"`, and `"HARTFORD"` are three different
strings. That has two practical consequences:

- **Filtering** silently misses rows. `filter(town == "hartford")` returns
  nothing if the data stores `"Hartford"`.
- **Joins** silently drop matches. If one table has `"New Haven"` and another
  has `"NEW HAVEN"`, those rows won't line up, and you may not notice the gap
  until your counts look wrong.

The defense is to normalize case *before* you filter or join, using
`stringr`'s case helpers (also from the
[Cleaning Data chapter](../02-cleaning-data/index.qmd)):

```{r}
#| eval: false
library(stringr)

str_to_lower("HARTFORD") # "hartford"  — good for matching
str_to_title("new haven") # "New Haven" — good for display
```

A common pattern is to lowercase both sides of a join key so the match doesn't
depend on how each source happened to capitalize things.

### Blanks, empties, and "missing"

An empty cell is more ambiguous than it looks, and different systems disagree on
what it means. When R reads a file, a "missing" value might arrive as any of:

| What's in the file | How R may see it | What it means |
|---|---|---|
| truly empty cell | `NA` | no value recorded |
| empty text `""` | `""` (empty string) | a value that happens to be blank text |
| the literal text `NA`, `N/A`, `.`, `Unknown` | a character string | a *placeholder* someone typed for missing |

That last row is the trap. If a source system writes `"N/A"` or `.` (a
convention SAS users will recognize) into blank cells, R reads them as ordinary
text, not as missing. A count of missing values will say zero, and any numeric
column with `"N/A"` sprinkled in will refuse to be numeric at all.

`read_csv()` has an `na =` argument for exactly this — you list the strings that
should be treated as missing:

```{r}
#| eval: false
library(readr)

# treat blanks, "NA", "N/A", and a lone "." all as missing
covid_data <- read_csv(
  csv_path,
  na = c("", "NA", "N/A", ".")
)
```

Why fuss over blank-versus-`NA`? Because it changes your numbers. An empty
string `""` is a real (if empty) value and gets **counted**; an `NA` does not.
And functions like `mean()` and `sum()` return `NA` the moment a single `NA` is
present unless you tell them to skip it with `na.rm = TRUE`. Getting the missing
values labeled correctly up front is what makes counts, denominators, and
`na.rm` behave the way you expect downstream. The
[Cleaning Data chapter](../02-cleaning-data/index.qmd) covers what to do with
`NA`s once they're correctly identified (`drop_na()`, `replace_na()`).

## This list will grow

These are the gotchas that come up again and again, across nearly every format
an epidemiologist reads. As the book picks up more formats — databases, APIs,
and other sources — new format-specific quirks will land here alongside them.
The reassuring part is that the *core* cleanup — dates, names, case, and
missing values — stays the same no matter how the data arrived.
