---
title: "Writing Data Files"
---

Once you have cleaned or summarized data, you will want to write it back out.
This chapter covers the most common output formats.

## CSV

`write_csv()` from `readr` is the default for analysis outputs that feed into
other scripts or tools.

```{r}
#| eval: false

library(readr)
library(here)

output_path <- here("data", "processed", "ct_covid_clean.csv")

write_csv(your_data, file = output_path)
```

## Checking and cleaning up with `fs`

`fs` lets you inspect and manage files from inside R — useful for verifying
that a write succeeded or cleaning up intermediate files.

```{r}
#| eval: false

library(fs)
library(here)

# list everything in a folder
dir_ls(path = here("data", "processed"))

# check whether specific files exist
files <- dir_ls(path = here("data"), glob = "*my-new-file*")
file_exists(files)

# delete files
file_delete(files)
file_exists(files)
```

## Parquet

Parquet is a compressed columnar format that reads and writes much faster
than CSV for large datasets. It is a better default than CSV once your
data exceeds a few hundred thousand rows.

```{r}
#| eval: false

install.packages("arrow")
```

```{r}
#| eval: false

library(arrow)
library(here)

output_path <- here("data", "processed", "ct_covid_clean.parquet")

write_parquet(your_data, sink = output_path)

# reading it back
your_data <- read_parquet(output_path)
```

Parquet files are not human-readable in a text editor, but they are
interoperable — Python, Spark, DuckDB, and most modern data tools can
read the same file.
