---
title: "Getting Ready for working with files"
---

## The Setup

Covid-19 challenged Public Health infrastructure in so many multi-faceted 
ways. The sheer amount of data collected and processed over the pandemic
meant that what worked in the beginning of the pandemic was no longer
feasible towards the end of the pandemic.

Over the course of this book we will explore Connecticut pandemic 
data and how to deal with the ever increasing amount of data. We will
start simple and do things manually before moving on to refine our 
analytics methodologies.

## Data Intro

We are going to use a snippet of real pandemic-era covid-19 data from the 
connecticut department of public health's open data portal^[[CT Open Data Portal](https://data.ct.gov/)].  
The full dataset can be found on the portal^[[Archived Covid Data](https://data.ct.gov/Health-and-Human-Services/COVID-19-Tests-Cases-and-Deaths-By-Town-ARCHIVE/28fr-iqnx/about_data)] 
but we will start simply and use the provided snippet in this  

TODO: add evergreen link 

repository `chapters/working-with-real-data/data/covid-2022.csv`

## Setup

R has many tools for working with data. Over the years the R community
has built up a number of their own tools and one of the most popular, 
the `tidyverse` will be our entrypoint into working with data in R.

R comes with a base set of tools and eager epidemiologists can opt to use
only the base suite of tool and go out into the world if they so choose.
However, many smart developers and epidemiologists have come before us
and we do not need to recreate the wheel. R has what are called
`packages' - or collection of ready made code that folks can use to
get work done.

TODO: -- talk about the CRAN etc etc....



We're going to slowly introduce the different parts of the tidyverse individually....

TODO: flesh 

### Installing Packages

```{r}
#| label: install-packages
#| eval: false 

# pulls the code down from the CRAN for folks to use and installs on
# their computer
install.packages(
  "dplyr", # for manipulating data
  "readr", # for reading common data files from a users' computers
  "fs", # for working with files on users' file systems
  "readxl", # for reading data from .xlsx files
  "writexl", # for writing data to .xlsx files
)
```

## Reading Data Files

### Reading our Comma Delimited File (.csv)

We are going to do a few things here:

1. 'Invoke' the packages we will be using so they are availabe to us.
1. Create and capture the path to our data file and save inside of an object
`csv_path`.
1. Read the data into our session from the csv and save inside of an object `covid_data`.

```{r}
#| label: setup-libraries

# Users can have all sorts of packages on their computer
# but will only need a limited subset of them at any one time
# add these packages to your session with `library()` function.

library(dplyr)
library(readr)
library(fs)

# we have a file - how do we read it in?
# we use fs's path() function to create a path
# we user readr's read_csv() and point it to  our path

csv_path <- path("data/covid-2022.csv")

covid_data <- read_csv(csv_path)
```

We have our data saved into our session, now what?
Let's take a look at it!

```{r}
# taking a look at the top few rows
head(covid_data)
```

```{r}
# glimpse from dplyr
glimpse(covid_data)
```


### Reading .xlsx files


We are going to use a new package `readxl` to read another
common data file type: '.xlsx' excel files. We will do
something similar to before but we will pick the excel file path.

```{r}
# our libraries above are still at play - no need to call again

library(readxl)

xlsx_path <- path("data/covid-2022.xlsx")
covid_data <- read_excel(xlsx_path)

covid_data
```

## Writing to files

### Writing to .csv

With our `covid_data` still in memory we are going to write it back
to our computer as a file.

It is easy to write .csv's user `readr`, .xlsx requires another package
`writexl`

```{r}

library(writexl)

# csv
new_csv_path <- path("data/my-new-file.csv")

write_csv(
  covid_data,
  file = new_csv_path
)


# xlsx
new_xlsx_path <- path("data/my-new-file.xlsx")

write_xlsx(
  covid_data,
  path = new_xlsx_path
)
```

Now we are going to use `fs` to check the contents of our data folder
and then clean up after ourselves.

```{r}

# check the data/ folder
dir_ls(
  path = path("data/")
)
```


The new files are there, but we already have that data.  We are going
to select *only* the file paths for our new files using `fs` and
then delete them from our space.


```{r}

files <-
  dir_ls(
    path = path("data/"),
    glob = "*my-new-file*" # use glob to search for text patterns
  )

file_exists(files)

file_delete(files)

file_exists(files)
```