---
title: "Loading Your First Dataset"
packages: [dplyr, readr, here]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Your workspace is set up and your raw data is sitting in `data/raw/`.
Time to write your first R code.

## Packages

R ships with a base set of tools. On top of that, the community has built
thousands of *packages* — collections of ready-made code you can install
and use without writing everything yourself. These are distributed through
**CRAN** (the Comprehensive R Archive Network), R's official package
repository.

The family of packages we'll lean on most is called the **tidyverse** —
a curated set of tools that share a common design and work well together.
We'll introduce them one at a time. This chapter uses two of them plus
one helper for building file paths.

### Installing Packages

```{r}
#| label: install-packages
#| eval: false

# Run this once to install the packages onto your computer.
install.packages(c(
  "dplyr", # for manipulating data
  "readr", # for reading common data files
  "here" # for building paths that work everywhere
))
```

## Reading Data Files

### Reading our Comma Delimited File (.csv)

We are going to do a few things here:

1. Load the packages we need into our session.
2. Build a path to our data file and save it as `csv_path`.
3. Read the data into our session and save it as `covid_data`.

```{r}
#| label: setup-libraries

library(dplyr)
library(readr)
library(here)
```

### Paths and `here()`

Before reading the file, a word on paths — because this is where R trips
people up early.

When you run code interactively, R looks for files starting from your project
root (wherever your `.Rproj` file lives or the root of the folder you opened
your with your IDE).
You'll often need to reference files that live somewhere other than the
project root. Hardcoding a full path like
`/home/my-username/folder-that-only-i-have/r-file.R` ties your project to
one specific computer. That's not ideal, which is why building paths
relative to the project root matters.

`here()` from the `here` package fixes both problems. It detects your project
root and always builds paths from there, regardless of how or where the code
runs. Reach for it any time you're building a path to a data file, instead of
pasting in a full path from your own machine or typing a relative path like
`"data/raw/file.csv"` that quietly breaks the moment your working directory
isn't where you assumed it was.

```{r}
#| eval: false

# here() takes folder names as separate arguments.
# It joins them with the correct separator for your OS.
csv_path <- here("data", "raw", "ct_covid_by_town_2026-06-30.csv")

covid_data <- read_csv(csv_path)
```

```{r}
#| include: false

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```

We have our data saved into our session — now let's look at it.

```{r}
# taking a look at the top few rows
head(covid_data)
```

```{r}
# glimpse from dplyr
glimpse(covid_data)
```


::: {.callout-note}
Working with Excel files (`.xlsx`) is covered in [Working with Excel Files](../further-reading/excel-files.qmd).
:::

::: {.callout-note}
Writing data to files — CSV, parquet, and others — is covered in
[Writing Data Files](../further-reading/writing-files.qmd).
:::

### Why build paths?

It's a small thing for one file, but the habit pays off fast: the same
`here()` pattern works whether you're reading one file or reading from
several folders across a project, and it keeps working when you hand
the project to a teammate or open it on a different machine.