---
title: "Loading Your First Dataset"
---

Your workspace is set up and your raw data is sitting in `data/raw/`.
Time to write your first R code.

## Packages

R ships with a base set of tools. On top of that, the community has built
thousands of *packages* — collections of ready-made code you can install
and use without writing everything yourself. These are distributed through
**CRAN** (the Comprehensive R Archive Network), R's official package
repository.

The family of packages we'll lean on most is called the **tidyverse** —
a curated set of tools that share a common design and work well together.
We'll introduce them one at a time. This chapter uses three of them plus
two helpers for reading and writing files.

### Installing Packages

```{r}
#| label: install-packages
#| eval: false

# Run this once to install the packages onto your computer.
install.packages(c(
  "dplyr",   # for manipulating data
  "readr",   # for reading common data files
  "fs",      # for working with files and paths
  "readxl",  # for reading .xlsx files
  "writexl"  # for writing .xlsx files
))
```

## Reading Data Files

### Reading our Comma Delimited File (.csv)

We are going to do a few things here:

1. 'Invoke' the packages we will be using so they are available to us.
2. Create and capture the path to our data file and save it inside an object called `csv_path`.
3. Read the data into our session from the csv and save it inside an object called `covid_data`.

```{r}
#| label: setup-libraries

# Users can have all sorts of packages on their computer
# but will only need a limited subset of them at any one time.
# Add these packages to your session with the `library()` function.

library(dplyr)
library(readr)
library(fs)

# We have a file - how do we read it in?
# Use fs's path() function to create a path,
# then readr's read_csv() to load it.

csv_path <- path("data/ct-covid-by-town-2026-06-30.csv")

covid_data <- read_csv(csv_path)
```

We have our data saved into our session, now what?
Let's take a look at it!

```{r}
# taking a look at the top few rows
head(covid_data)
```

```{r}
# glimpse from dplyr
glimpse(covid_data)
```


### Reading .xlsx files

Many files you'll encounter in public health come as Excel spreadsheets.
We use `readxl` to read them the same way we read CSVs.

```{r}
#| eval: false

# our libraries above are still at play — no need to call them again

library(readxl)

xlsx_path <- path("data/your-file.xlsx")
covid_data <- read_excel(xlsx_path)

covid_data
```

## Writing to files

### Writing to .csv

With our `covid_data` still in memory we are going to write it back
to our computer as a file.

Writing .csv files uses `readr`; writing .xlsx requires the `writexl` package.

```{r}

library(writexl)

# csv
new_csv_path <- path("data/my-new-file.csv")

write_csv(
  covid_data,
  file = new_csv_path
)


# xlsx
new_xlsx_path <- path("data/my-new-file.xlsx")

write_xlsx(
  covid_data,
  path = new_xlsx_path
)
```

Now we are going to use `fs` to check the contents of our data folder
and then clean up after ourselves.

```{r}

# check the data/ folder
dir_ls(
  path = path("data/")
)
```

The new files are there, but we already have that data. We are going
to select *only* the file paths for our new files using `fs` and
then delete them from our space.

```{r}

files <-
  dir_ls(
    path = path("data/"),
    glob = "*my-new-file*" # use glob to search for text patterns
  )

file_exists(files)

file_delete(files)

file_exists(files)
```
