---
title: "Loading Your First Dataset"
packages: [dplyr, readr, fs, here]
---

```{r}
#| label: chapter-root
#| include: false
library(here)
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

Your workspace is set up and your raw data is sitting in `data/raw/`.
Time to write your first R code.

## Packages

R ships with a base set of tools. On top of that, the community has built
thousands of *packages* — collections of ready-made code you can install
and use without writing everything yourself. These are distributed through
**CRAN** (the Comprehensive R Archive Network), R's official package
repository.

The family of packages we'll lean on most is called the **tidyverse** —
a curated set of tools that share a common design and work well together.
We'll introduce them one at a time. This chapter uses three of them plus
a few helpers for reading, writing, and navigating files.

### Installing Packages

```{r}
#| label: install-packages
#| eval: false

# Run this once to install the packages onto your computer.
install.packages(c(
  "dplyr", # for manipulating data
  "readr", # for reading common data files
  "fs", # for working with files and paths
  "here" # for building paths that work everywhere
))
```

## Reading Data Files

### Reading our Comma Delimited File (.csv)

We are going to do a few things here:

1. Load the packages we need into our session.
2. Build a path to our data file and save it as `csv_path`.
3. Read the data into our session and save it as `covid_data`.

```{r}
#| label: setup-libraries

library(dplyr)
library(readr)
library(fs)
library(here)
```

### Paths and `here()`

Before reading the file, a word on paths — because this is where R trips
people up early.

When you run code interactively, R looks for files starting from your project
root (wherever your `.Rproj` file lives or the root of the folder you opened
your with your IDE).
Often you will users will find themselves referencing files in various places
that are not in the root of their project.
Hardcoding out a file path like `/home/my-username/folder-that-only-i-have/r-file.R`
makes this project tied only to that user's computer. That is not ideal,
which is why using paths relative to the project root  is so important.

`here()` from the `here` package fixes both problems. It detects your project
root and always builds paths from there, regardless of how or where the code
runs. Use it instead of `path()` whenever you are building paths to data files.

```{r}
#| eval: false

# here() takes folder names as separate arguments.
# It joins them with the correct separator for your OS.
csv_path <- here("data", "raw", "ct_covid_by_town_2026-06-30.csv")

covid_data <- read_csv(csv_path)
```

```{r}
#| include: false

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```

We have our data saved into our session — now let's look at it.

```{r}
# taking a look at the top few rows
head(covid_data)
```

```{r}
# glimpse from dplyr
glimpse(covid_data)
```


::: {.callout-note}
Working with Excel files (`.xlsx`) is covered in [Working with Excel Files](../further-reading/excel-files.qmd).
:::

::: {.callout-note}
Writing data to files — CSV, parquet, and others — is covered in
[Writing Data Files](../further-reading/writing-files.qmd).
:::

### Why build Paths?

Building paths makes your code more more robust and shareable with others -
but why go through the trouble for a single path? Setting aside what
we discussed earlier about hardcoding paths, this is a good habit
to build because these paths can become a springboard for more advanced
workflows. Programming analytic workflows is often about building small 
habits and muscle memory that eventually compound into rigourous analytic 
practices that can then be used elsewhere. That path you built? Now that
can become a lever you can pull to read in multiple files all across a
project, or even across multiple projects. 

This small habit is about more than a path, it is about breaking down a 
problem into smaller pieces, and leaving little bread crumbs for future you 
and teammates. The goal may be the insights at the end of your analysis, but 
if done well, you can take your analysis on the road and offer up insights
inside of your organization and beyond.  This is the power of reproducible
research and analytics.