---
title: "A First Look at Hartford"
---

```{r}
#| label: project-root
#| include: false
library(here)
source(here::here("preamble-data.R"))
```

It's Monday. The incident-command call is at 4 p.m., and the
Commissioner's office wants the Hartford numbers in writing before it —
the first standing question from chapter 0: **is it getting worse in
Hartford?**

The data is loaded. You have until 4 o'clock to turn it into an answer.

## Setup

```{r}
#| label: setup
#| eval: false

library(dplyr)
library(readr)
library(here)

covid_data <- read_csv(here("data", "raw", "ct_covid_by_town_2026-06-30.csv"))
```

```{r}
#| label: setup-exec
#| include: false

library(dplyr)
library(readr)
library(here)

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")
```

## The pipe

Before diving in, one piece of syntax you'll see throughout this book:

```{r}
#| eval: false

# The pipe |> passes the result of one step into the next.
# Read it as "and then."
covid_data |>
  filter(Town == "Hartford")

# is the same as:
filter(covid_data, Town == "Hartford")
```

Coming from SAS, think of it like chaining `DATA` steps and `PROC` steps
without writing an intermediate dataset between each one.

::: {.callout-tip}
## Keyboard shortcuts

| Shortcut | What it does |
|---|---|
| `Ctrl+Shift+M` | Insert the native pipe, once you turn that option on |
| `Ctrl+Enter` | Run the current line or selection |
| `Ctrl+Shift+Enter` | Run the entire current chunk |

Out of the box that shortcut inserts the older `%>%` pipe. Flip it in
Tools → Global Options → Code → Editing → "Use native pipe operator" and
you'll get `|>` like the rest of this book.

Learn all three early — you will use them on every line you write.
:::

## Step 1 — Filter to Hartford

`filter()` keeps only the rows that match a condition.

```{r}
hartford <-
  covid_data |>
  filter(Town == "Hartford")

nrow(hartford)
```

## Step 2 — Select and rename columns

`glimpse()` shows that some column names have spaces — a common headache
with real-world data. `select()` lets us pick the columns we want and
rename them at the same time.

Column names with spaces need backticks when you reference them:

```{r}
hartford <-
  hartford |>
  select(
    date = `Last update date`,
    town = Town,
    total_cases = `Total cases`,
    total_deaths = `Total deaths`
  )

glimpse(hartford)
```

## Step 3 — Parse the date

The date column came in as a character string. We need it as an actual
date so we can sort and eventually plot it correctly.

`mutate()` adds or modifies columns. `parse_date()` from `readr` converts
a character string to a date given a format string.

```{r}
hartford <-
  hartford |>
  mutate(
    date = parse_date(
      date,
      format = "%m/%d/%Y"
    )
  )

glimpse(hartford)
```

## Step 4 — Sort by date

`arrange()` sorts rows. The data comes in reverse-chronological order;
let's flip it so the earliest dates come first.

```{r}
hartford <-
  hartford |>
  arrange(date)

head(hartford)
```

## Step 5 — Computing daily new cases

`total_cases` is a running total — it only ever goes up. To see whether
things are getting *worse* day to day, we need the change between one
report and the next, not the running total itself. `lag()` shifts a
column down by one row, so `total_cases - lag(total_cases)` gives
"today's total minus yesterday's total" — the daily new cases.

```{r}
hartford <-
  hartford |>
  mutate(new_cases = total_cases - lag(total_cases))

tail(hartford, 14)
```

The very first row has no "yesterday" to subtract, so `lag()` gives it
`NA` there — expected, not a bug.

::: {.callout-warning}
## Handle "as reported" data with care

`new_cases` here is only as good as the reporting behind it. Real
pandemic data was reported in batches — weekends and holidays often
showed artificially low counts followed by a backlog dump on the next
business day, and health departments regularly went back and revised
earlier totals as duplicate tests were removed or late results came in.
That means `new_cases` can come out *negative* on a day when a
correction lowered a previous total, even though nothing was actually
"un-reported." This wasn't a quirk of this dataset specifically — it
was one of the most persistent, genuinely frustrating problems in
tracking COVID-19 nationally. It's worth remembering any time "new
cases" is computed by subtracting two "as of" snapshots instead of
coming from a system that recorded each case on the day it actually
happened.
:::

## Putting it together

Each step above modified `hartford` and saved it back. We can write that
entire chain as one pipeline:

```{r}
#| eval: false

hartford <-
  covid_data |>
  filter(Town == "Hartford") |>
  select(
    date = `Last update date`,
    town = Town,
    total_cases = `Total cases`,
    total_deaths = `Total deaths`
  ) |>
  mutate(date = parse_date(date, format = "%m/%d/%Y")) |>
  arrange(date) |>
  mutate(new_cases = total_cases - lag(total_cases))
```

## What does the data say?

```{r}
tail(hartford, 14)
```

Two weeks of Hartford data, with daily new cases and running totals.
This is the foundation the briefing will be built on — clean, sorted,
and scoped to the question.

Note that `total_cases` is cumulative: it only goes up. `new_cases` is
the day-over-day change we just computed — that's the number to watch
when asking whether things are getting worse, with the reporting
caveat above kept in mind.

## The answer for the 4 p.m. call

A table isn't an answer. The Commissioner's office asked a yes-or-no
question, so give them one. Compare the most recent week of reports
against the week before — a week's average rides over the weekend dips
and one-day corrections that a single date would trip on:

```{r}
#| label: finding

recent <- hartford |> filter(!is.na(new_cases))

last_week <- mean(tail(recent$new_cases, 7))
prior_week <- mean(head(tail(recent$new_cases, 14), 7))

verdict <- if (last_week > prior_week) "getting worse" else "holding steady or improving"

verdict
```

That's the line for this week's Situation Brief: **Hartford is
`r verdict`** — its daily new cases averaged about `r round(last_week, 1)`
over the last week of reports, against `r round(prior_week, 1)` the week
before. One sentence, backed by the two-week table above, with the
reporting caveat noted so nobody over-reads a single day.

That's the whole difference between a query and a briefing: a briefing
ends in an answer, not a data dump. Every chapter from here adds to how
that answer is built — but the shape never changes.

## Saving our data artifact

Imagine this Hartford data was the end goal — now would be the time
to save the data in memory to a file in our `data/processed/` folder.

```{r}
# write_csv(...)
```

::: {.callout-note}
Want to compare against a reference? See
[`project-snapshots/after-ch-01`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/after-ch-01)
on GitHub for the `scripts/` this chapter should leave you with.
:::
