---
title: "A First Look at Hartford"
---

```{r}
#| label: project-root
#| include: false
library(here)
.chapter_data <- here::here("chapters", "01-working-with-real-data", "data")
```

The scenario question from chapter 0: **Is it getting worse in Hartford?**

The data is loaded. Now let's answer it.

## Setup

```{r}
#| label: setup
#| eval: false

library(dplyr)
library(readr)
library(here)

covid_data <- read_csv(here("data", "raw", "ct_covid_by_town_2026-06-30.csv"))
```

```{r}
#| label: setup-exec
#| include: false

library(dplyr)
library(readr)
library(fs)
library(here)

covid_data <- read_csv(fs::path(
  .chapter_data,
  "ct-covid-by-town-2026-06-30.csv"
))
```

## The pipe

Before diving in, one piece of syntax you'll see throughout this book:

```{r}
#| eval: false

# The pipe |> passes the result of one step into the next.
# Read it as "and then."
covid_data |>
  filter(Town == "Hartford")

# is the same as:
filter(covid_data, Town == "Hartford")
```

Coming from SAS, think of it like chaining `DATA` steps and `PROC` steps
without writing an intermediate dataset between each one.

::: {.callout-tip}
## Keyboard shortcuts

| Shortcut | What it does |
|---|---|
| `Ctrl+Shift+M` | Insert the pipe `|>` |
| `Ctrl+Enter` | Run the current line or selection |
| `Ctrl+Shift+Enter` | Run the entire current chunk |

Learn all three early — you will use them on every line you write.
:::

## Step 1 — Filter to Hartford

`filter()` keeps only the rows that match a condition.

```{r}
hartford <-
  covid_data |>
  filter(Town == "Hartford")

nrow(hartford)
```

## Step 2 — Select and rename columns

`glimpse()` shows that some column names have spaces — a common headache
with real-world data. `select()` lets us pick the columns we want and
rename them at the same time.

Column names with spaces need backticks when you reference them:

```{r}
hartford <-
  hartford |>
  select(
    date = `Last update date`,
    town = Town,
    new_positives = `Number of positives`,
    total_cases = `Total cases`,
    total_deaths = `Total deaths`
  )

glimpse(hartford)
```

## Step 3 — Parse the date

The date column came in as a character string. We need it as an actual
date so we can sort and eventually plot it correctly.

`mutate()` adds or modifies columns. `parse_date()` from `readr` converts
a character string to a date given a format string.

```{r}
hartford <-
  hartford |>
  mutate(
    date = parse_date(
      date,
      format = "%m/%d/%Y"
    )
  )

glimpse(hartford)
```

## Step 4 — Sort by date

`arrange()` sorts rows. The data comes in reverse-chronological order;
let's flip it so the earliest dates come first.

```{r}
hartford <-
  hartford |>
  arrange(date)

head(hartford)
```

## Putting it together

Each step above modified `hartford` and saved it back. We can write that
entire chain as one pipeline:

```{r}
#| eval: false

hartford <-
  covid_data |>
  filter(Town == "Hartford") |>
  select(
    date = `Last update date`,
    town = Town,
    new_positives = `Number of positives`,
    total_cases = `Total cases`,
    total_deaths = `Total deaths`
  ) |>
  mutate(date = parse_date(date, format = "%m/%d/%Y")) |>
  arrange(date)
```

## What does the data say?

```{r}
tail(hartford, 14)
```

Two weeks of Hartford data, with new daily positives and running totals.
This is the foundation the briefing will be built on — clean, sorted,
and scoped to the question.

Note that `total_cases` is cumulative: it only goes up. `new_positives`
is the count of new positive tests reported for that date — that is the
number to watch when asking whether things are getting worse.

## Saving our data artifact

Imagine that these hartford data was the end goal - now would be the time
to save the data in memory to a file in our `data/processed/` folder.

```{r}
# write_csv(...)
```
