---
title: "Visualizing Your Data"
---

```{r}
#| label: setup-themes
#| include: false
source(here::here("preamble-themes.R"))
```

```{r}
#| label: chapter-root
#| include: false
library(here)
library(readr)
library(dplyr)
library(ggplot2)
library(sf)

.chapter_data <- here::here("chapters", "03-data-visualization", "data")

covid_data <- read_csv(fs::path(
  .chapter_data,
  "ct-covid-by-town-2026-06-30.csv"
))

hartford <-
  covid_data |>
  filter(Town == "Hartford") |>
  select(
    date = `Last update date`,
    new_positives = `Number of positives`,
    total_cases = `Total cases`,
    total_deaths = `Total deaths`
  ) |>
  mutate(date = parse_date(date, format = "%m/%d/%Y")) |>
  arrange(date)
```

You've answered the question with numbers. Now it's time to make it
visible.

In Chapter 1 you filtered Hartford's data down to a single tidy tibble.
You looked at the rows in your console. Numbers tell the truth, but they
don't make it easy to see at a glance **how** the case counts moved over
those two weeks.

This chapter teaches you how to turn that tibble into a plot that tells
the story to someone else — the Commissioner, a journalist, a colleague
who doesn't want to open a CSV.

The tool is `ggplot2`, and it works very differently from the `plot()`
function you might have used before. That's good. `plot()` gives you
a blank canvas and makes you draw every line. `ggplot2` gives you
a grammar — a language for describing *what* you want the plot to show
—and it handles the drawing.

## Installing `ggplot2`

Before this chapter can work, install the package once.

```{r}
#| eval: false

install.packages("ggplot2")

# You'll also install extra packages later in this chapter for
# mapping and choropleths. They're small and useful.
install.packages(c("tigris", "sf", "mapview"))
```

Load `ggplot2` along with the packages you've used so far:

```{r}
#| label: setup-libs
library(ggplot2)
```

## Building a plot layer by layer

### The Grammar of Graphics

`ggplot2` is built on a principle called the **Grammar of Graphics**.
It says every plot is made of layers, each adding something to the
picture:

| Layer | What it does | Example |
|-------|-------------|---------|
| **Data** | The tibble to plot | `hartford` |
| **Aesthetic mapping** | Which columns go on which axis, or to color, size, etc. | `x = date, y = new_positives` |
| **Geometric object** | The shape of the marks | dots, lines, bars, areas |
| **Theme** | Non-data ink: fonts, backgrounds, axis styling | removal of grid lines, Poppins font |

Think of it like building a sandwich. Data is the bread, aesthetics are
the spread, geometry is the meat, theme is the wrap. You can leave out
any layer and R will still give you something — but the layers only make
sense when they work together.

### Your first plot: a scatter plot

We'll start with the simplest possible scatter plot — dots — because it
teaches the most.

```{r}
#| label: intro-scatter-simple
ggplot(
  data = hartford,
  mapping = aes(
    x = date,
    y = new_positives
  )
) +
  geom_point()
```

That's it. Three lines.

Let's read them back:

1. `ggplot()` starts the plot, supplies the data and the *aesthetic
   mapping* (`aes`) that says "`date` goes on the x-axis and
   `new_positives` goes on the y-axis".
2. `+ geom_point()` adds the layer of dots. The `+` is what joins
   layers in `ggplot2` — always with a space before and after the `+`.
3. Every geometry layer inherits the data and aesthetics from `ggplot()`
   unless you override them inside `aes()` for that specific layer.

The default ggplot2 output works well for internal work, but a briefing
needs something cleaner. Let's build on that first plot.

### From dots to a line

Our data is time-series: `date` on the x-axis, values on the y-axis.
A line makes the trend easier to read than dots alone:

```{r}
#| label: intro-line
ggplot(
  data = hartford,
  mapping = aes(
    x = date,
    y = new_positives
  )
) +
  geom_point(alpha = 0.1) +
  geom_line()
```

Two geometry layers on one plot — `geom_point()` and `geom_line()`.
Both inherited the data and aesthetic mapping from `ggplot()`, so we
never repeated them.

The plot is starting to look like what the Commissioner wants. But it
has the default theme's gray background, the default axis labels read like
column names, and the Poppins font isn't there yet. Let's fix each of
those in turn.

### Title, axis labels, and theme

`labs()` replaces the automatic labels from column names and adds a
title. `theme_ct_vertical()` (defined in `preamble.Rmd`) removes the
y-axis line and tick marks for a cleaner look.

```{r}
#| label: intro-annotated
ggplot(
  data = hartford,
  mapping = aes(
    x = date,
    y = new_positives
  )
) +
  geom_point() +
  geom_line() +
  labs(
    title = "New COVID-19 Cases — Hartford",
    subtitle = "Daily count from report to report (Jun–Jul 2026)",
    x = NULL,
    y = "New Cases on Each Date"
  ) +
  theme_ct_vertical()
```

`x = NULL` removes the x-axis label entirely — the title is enough, and
everyone reading this knows it shows dates.



### Adding color

The default plot line uses the first color in the ggplot2 palette.
In public health work you probably want a specific color — CT DPH blue,
or a color that's immediately readable in color-blind mode.

Map the `color` aesthetic to a string (a color name) to tint both points
and line:

```{r}
#| label: intro-colored
ggplot(data = hartford, mapping = aes(x = date, y = new_positives)) +
  geom_point(color = "#3371E7", size = 2, alpha = 0.1) +
  geom_line(color = "#3371E7", linewidth = 1) +
  labs(
    title = "New COVID-19 Cases — Hartford",
    subtitle = "Daily count from report to report (Jun–Jul 2026)",
    x = NULL,
    y = "New Cases on Each Date"
  ) +
  theme_ct_vertical() +
  theme_poppins_bold_all() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50")
  )
```

Two things to notice:

1. `color = "#3371E7"` is set **outside** `aes()` because it's a fixed
   value, not a column. Inside `aes()`, R would try to map the literal
   word "#3371E7" to a legend. You rarely want that.
2. `theme_poppins_bold_all()` brings the Poppins font everywhere, but
   we still tweak `plot.title` and `plot.subtitle` separately for sizes.
   Theme layers stack — later ones fill gaps left by earlier ones.

That last plot is the kind of thing you'd paste directly into a briefing
slide. You built it piece by piece, and each piece does one thing.

### Saving the plot

When you're happy with a plot, save it to `outputs/figures/`:

```{r}
#| label: save-plot-hartford
#| eval: false 
ggsave(
  here("outputs", "figures", "hartford_new_cases.png"),
  width = 7,
  height = 4,
  dpi = 300
)
```

`ggsave()` saves the last plot you rendered. No need to name it
explicitly — just call it immediately after running the plot.

## Aesthetic mapping

So far you've been setting the position of the plot (x and y axis).
`ggplot2` also maps aesthetics to *other* properties:

| Aesthetic | What it controls | Accepts |
|-----------|-----------------|---------|
| `x`, `y` | Position on the axes | numeric, date, character |
| `color` | Stroke/border color of points and line | column name or fixed color |
| `fill` | Fill color of bars, areas | column name or fixed color |
| `size` | Size of points or line thickness | column name or fixed size |
| `shape` | Point shape (0–24 different shapes) | column name or fixed shape |
| `alpha` | Transparency (0 = invisible, 1 = opaque) | column name or fixed value |
| `linetype` | Dashed, dotted, etc. | column name or fixed value |

When you put a **column name** inside `aes()`, ggplot2 creates a legend
that tells the reader which values map to which colors, sizes, shapes
etc. When you set one **outside** `aes()`, the value is fixed and
no legend is generated.

To see it in action, add a second line to the same plot — one for
`total_cases` (cumulative) — by mapping `y` to a column:

```{r}
#| label: aesthetic-mapping
ggplot(data = hartford, mapping = aes(x = date)) +
  geom_line(aes(y = new_positives, color = "New Cases"), linewidth = 1) +
  geom_line(aes(y = total_cases, color = "Cumulative Cases"), linewidth = 1) +
  scale_color_manual(
    name = "Metric",
    values = c("New Cases" = "#F27124", "Cumulative Cases" = "#3371E7")
  ) +
  labs(
    title = "New Cases vs. Cumulative Cases — Hartford",
    y = "Count",
    color = "Metric"
  ) +
  theme_ct_vertical() +
  theme_poppins_bold_all()
```

Because both `new_positives` and `total_cases` are columns with the same
meaning (both are counts), you could also reshape the data to long
format so you don't repeat `geom_line()` at all. We'll cover `pivot_longer()`
in a later chapter. For now, this is a fast way to show the same plot
with two series.

## Choropleth maps

A scatter plot shows change over time. A map shows change across
space. For a state-level briefing, the Commissioner might want to see
case counts *across all the towns at once*.

The `tigris` package fetches official boundary files from the Census
Bureau. In our case, we want town boundaries for Connecticut:

```{r}
#| label: choropleth-choro

library(tigris)
library(sf)

# Fetch CT town boundaries
ct_towns <-
  county_subdivisions(
    state = "09",
    cb = TRUE,
    progress_bar = FALSE
  )

# Show just the outlines on a blank canvas
ggplot() +
  geom_sf(data = ct_towns, fill = "gray95", color = "black", linewidth = 0.1) +
  theme_void()
```

That's Connecticut's 169 towns — just outlines. `geom_sf()` draws
spatial data (the `sf` geometry column). The rest follows the same
pattern as every other ggplot2 plot: `ggplot() + geom_sf() + labs() +
theme()`.

### Coloring towns by case rate

Now join the COVID data to the map by town name and map `fill` to the
case rate:

```{r}
#| label: choropleth-data

# Summarize cases by town on the latest date
latest_date <- max(covid_data$`Last update date`, na.rm = TRUE)

town_cases <-
  covid_data |>
  filter(`Last update date` == latest_date) |>
  select(
    Town = Town,
    case_rate = `Case rate`,
    total_cases = `Total cases`
  ) |>
  rename(name = Town)

# Join COVID data to spatial data
ct_bounded <- ct_towns |>
  left_join(town_cases, by = join_by(NAME == name))
```

::: {.callout-tip}
## What just happened: `left_join` explained

`left_join(x, y, by)` keeps every row from the **left** table (`x`) and
adds matching columns from the **right** table (`y`). Where there's no
match, the new columns are filled with `NA`. Think of it as: "give me
all the towns, and bring in any COVID numbers that line up."

We cover joins in detail in a later chapter — [see Further
Reading](further-reading/index.qmd) for the full guide.
:::

```{r}
#| label: choropleth-fill

# Choropleth: gray for towns without data, blue gradient for those with
ggplot() +
  geom_sf(
    data = ct_bounded[is.na(ct_bounded$case_rate), ],
    fill = "gray95",
    color = "white",
    linewidth = 0.1
  ) +
  geom_sf(
    data = ct_bounded[!is.na(ct_bounded$case_rate), ],
    mapping = aes(fill = case_rate),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_gradient2(
    low = "#C6D4FB",
    mid = "#7094F5",
    high = "#00214D",
    midpoint = mean(ct_bounded$case_rate, na.rm = TRUE),
    na.value = "gray95",
    name = "Case Rate\n(per 100,000)"
  ) +
  labs(
    title = "COVID-19 Case Rate by Town",
    subtitle = paste("As of", latest_date),
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    legend.position = "right"
  )

The choropleth tells at a glance which towns carry the heaviest burden.
Towns without data (suppressed for small sample sizes) appear gray.

::: {.callout-note}
`geom_sf()` replaces `geom_point()`, `geom_line()`, and others when
your data is spatial. It reads the geometry column automatically and
draws polygons, lines, or points based on the geometry type.
:::

::: {.callout-note}
This chapter gave you the foundations of `ggplot2` and a taste of
spatial visualization. The next chapters add tables (`gt`), Quarto
report-building, functions, and API fetching — the full arc from raw
data to self-updating report.
:::
