---
title: "Interactive Maps: mapview & leaflet"
packages: [mapview, leaflet]
---

```{r}
#| label: ml-setup
#| include: false
source(here::here("preamble-themes.R"))
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))

library(here)
library(readr)
library(dplyr)
library(sf)
library(tigris)
library(leaflet)
library(mapview)

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")

# Fetch CT town boundaries
ct_towns <-
  county_subdivisions(
    state = "09",
    cb = TRUE,
    progress_bar = FALSE
  )

# Summarize cases by town on the latest date
latest_date <- max(covid_data$`Last update date`, na.rm = TRUE)

town_cases <-
  covid_data |>
  filter(`Last update date` == latest_date) |>
  select(
    Town = Town,
    case_rate = `Case rate`,
    total_cases = `Total cases`
  ) |>
  rename(name = Town)

# Join COVID data to spatial data
ct_bounded <- ct_towns |>
  left_join(town_cases, by = join_by(NAME == name))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

The choropleth in the last section is built for a slide — you export it
once, drop it into a deck, and it doesn't change again until the next
briefing. That's exactly what a static `ggplot2` map is for.

Not every audience gets a slide, though. A web-facing report, an
internal dashboard, or an email to a colleague who wants to click around
the map themselves calls for something the reader can zoom, pan, and
hover over. That's an **interactive map**, and R gives you two ways to
build one: `mapview` for a fast look, and `leaflet` for a map you build
and control piece by piece.

Install both once:

```{r}
#| eval: false
install.packages(c("leaflet", "mapview"))
```

## The fast way: `mapview`

```{r}
#| label: mapview-quick
mapview(ct_bounded, zcol = "case_rate")
```

One line. `mapview()` looks at `ct_bounded`, sees it's spatial data,
and automatically:

- picks a basemap and drops your polygons on top of it
- colors each town by the `case_rate` column, since you named it in `zcol`
- builds a legend for that column
- wires up hover and click popups showing every attribute column

That's the trade-off: zero setup, but you don't control any of it.
Use `mapview()` while you're exploring a dataset at your desk — checking
that a join worked, spotting an outlier town, eyeballing a spatial
pattern. When you need a specific popup, a specific color ramp, or a
specific basemap for a report someone else will read, reach for
`leaflet` instead.

## The deliberate way: `leaflet`

`leaflet` builds a map the same way `ggplot2` builds a plot — one piece
at a time. Where `ggplot2` chains layers with `+`, `leaflet` chains them
with the pipe. Read a `leaflet` chain the same way you read a `ggplot2`
chain: each step adds one thing to the map.

```{r}
#| label: leaflet-basic
leaflet(ct_bounded) |>
  addTiles() |>
  addPolygons()
```

Three steps:

1. `leaflet(ct_bounded)` starts the map and supplies the data — like
   `ggplot(data = ...)`.
2. `addTiles()` adds the basemap (streets, place names) from
   OpenStreetMap.
3. `addPolygons()` draws the town boundaries on top of it.

Right now every town is the same flat blue and tells you nothing. Let's
add the pieces one at a time: popups, then color.

### Adding popups

A popup is text that appears when someone clicks (or hovers over, with
`highlightOptions`) a polygon. Build the text yourself with `paste0()`,
one string per town, then pass the whole vector to `addPolygons()`:

```{r}
#| label: leaflet-popups
town_popup <- paste0(
  "<strong>",
  ct_bounded$NAME,
  "</strong><br/>",
  "Case rate: ",
  round(ct_bounded$case_rate, 1),
  " per 100,000<br/>",
  "Total cases: ",
  ct_bounded$total_cases
)

leaflet(ct_bounded) |>
  addTiles() |>
  addPolygons(popup = town_popup)
```

`town_popup` is a plain character vector, one entry per row of
`ct_bounded` — same length, same order. `<strong>` and `<br/>` are HTML
tags; `leaflet` popups render as HTML, so you can bold text and break
lines the same way you would on a web page.

### Coloring by case rate

`colorBin()` turns a numeric column into a small palette function —
give it bin breaks and colors, and it hands back a function that maps
any value in that range to a color. Build the ramp from the CT DPH
primary palette so the map matches every other plot in this book:

```{r}
#| label: leaflet-color
case_rate_palette <- colorBin(
  palette = c(
    list_ct_primary_colors$primary_shade4,
    list_ct_primary_colors$primary_shade3,
    list_ct_primary_colors$primary,
    list_ct_primary_colors$primary_shade2,
    list_ct_primary_colors$primary_shade1
  ),
  domain = ct_bounded$case_rate,
  bins = 5,
  na.color = "gray95"
)

ct_case_map <-
  leaflet(ct_bounded) |>
  addTiles() |>
  addPolygons(
    fillColor = ~ case_rate_palette(case_rate),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    popup = town_popup
  ) |>
  addLegend(
    pal = case_rate_palette,
    values = ~case_rate,
    title = "Case Rate<br/>(per 100,000)",
    position = "bottomright"
  )

ct_case_map
```

`primary_shade4` (the lightest blue) anchors the low end of the ramp,
`primary_shade1` (the darkest) anchors the high end — light means
fewer cases, dark means more, same logic as `scale_fill_gradient2()`
in the static choropleth. `~case_rate_palette(case_rate)` uses `leaflet`'s
formula syntax to say "run this function on the `case_rate` column of
the data I gave `leaflet()` at the start" — it plays the same role
`aes(fill = case_rate)` played earlier.

::: {.callout-tip}
Because we built `case_rate_palette` from `list_ct_primary_colors`
instead of hardcoded hex codes, if the CT DPH brand colors ever change,
updating `preamble-themes.R` updates this map too.
:::

## Marking a single town

You've already met Hartford — [the daily new-case line chart earlier in
this chapter](index.qmd#from-dots-to-a-line) walked through its case
counts day by day. A choropleth shows every town at once, but sometimes
you want to point a reader at one town in particular. `addMarkers()`
drops a pin at a specific coordinate on top of the map you already
built:

```{r}
#| label: hartford-marker
ct_case_map |>
  addMarkers(
    lng = -72.6734,
    lat = 41.7658,
    popup = "Hartford — click here to see the daily new-case trend for this town."
  )
```

Piping `addMarkers()` onto `ct_case_map` adds the pin without rebuilding
the choropleth underneath it — the same "keep adding layers" habit
you've been using since the first `ggplot2` scatter plot.

## Saving and exporting

An interactive map only stays interactive where something can run
JavaScript — a rendered HTML report, a Shiny app, a browser. Paste it
into a Word document or a PDF and it becomes a flat image anyway, so
`mapview::mapshot()` lets you export either format depending on where
the map is going.

Save the fully interactive version as its own HTML file — useful when
you want to embed it in a Quarto HTML report with an `<iframe>`, or
email it to someone who can open it and click around themselves:

```{r}
#| label: mapshot-html
#| eval: false
mapshot(
  ct_case_map,
  url = here(
    "chapters",
    "03-data-visualization",
    "maps",
    "ct-case-rate-map.html"
  )
)
```

Save a static snapshot as PNG when the destination can't run
JavaScript at all — a slide deck, a printed briefing, a PDF report:

```{r}
#| label: mapshot-png
#| eval: false
mapshot(
  ct_case_map,
  file = here(
    "chapters",
    "03-data-visualization",
    "maps",
    "ct-case-rate-map.png"
  )
)
```

Rule of thumb: if the report renders to HTML and the reader is looking
at a screen, keep the map interactive. If it renders to PDF, Word, or
a printed page, export a PNG instead.

::: {.callout-note}
This section covered enough `leaflet` to build a colored, clickable
choropleth. `leaflet` can also layer multiple datasets, add custom
basemaps, cluster hundreds of markers, and respond to clicks with
Shiny — Further Reading has a deeper dive into `leaflet` waiting for
when you need it.
:::
