---
title: "Building This Week's Briefing Deck"
---

```{r}
#| label: project-root
#| include: false
library(here)
source(here::here("preamble-themes.R"))
source(here::here("preamble-data.R"))
```

Chapter 2 left you with `covid_clean` — tidy snake_case names, real
`Date` values, trimmed and title-cased town names. Time to put that to
work. This week's Situation Brief needs two pictures: a trend chart
comparing Hartford to the rest of the state, and a statewide map. Both
have to land as files in `outputs/figures/` — which is exactly where the
next chapter's brief will go looking for them when it assembles the page.
You're not making slides here; you're stocking the brief.

## Setup

```{r}
#| eval: false

library(dplyr)
library(readr)
library(here)
library(janitor)
library(stringr)
library(lubridate)
library(ggplot2)
library(sf)
library(tigris)

covid_data <- read_csv(here("data", "raw", "ct_covid_by_town_2026-06-30.csv"))

covid_clean <-
  covid_data |>
  clean_names() |>
  mutate(
    last_update_date = mdy(last_update_date),
    town = town |> str_trim() |> str_to_title()
  ) |>
  arrange(town, last_update_date) |>
  group_by(town) |>
  mutate(new_cases = total_cases - lag(total_cases)) |>
  ungroup()
```

```{r}
#| label: setup-exec
#| include: false

library(dplyr)
library(readr)
library(here)
library(janitor)
library(stringr)
library(lubridate)
library(ggplot2)
library(sf)
library(tigris)

covid_data <- read_book_data("ct-covid-by-town-2026-06-30.csv")

covid_clean <-
  covid_data |>
  clean_names() |>
  mutate(
    last_update_date = mdy(last_update_date),
    town = town |> str_trim() |> str_to_title()
  ) |>
  arrange(town, last_update_date) |>
  group_by(town) |>
  mutate(new_cases = total_cases - lag(total_cases)) |>
  ungroup()
```

## Step 1 — Hartford vs. the statewide average

A single town's trend only means something next to a baseline. Build
Hartford's daily new-case series, then a statewide daily average, and
stack them into one tibble so both can be mapped to the same `color`
aesthetic.

```{r}
#| label: trend-data
hartford_trend <-
  covid_clean |>
  filter(town == "Hartford") |>
  select(last_update_date, new_cases) |>
  mutate(series = "Hartford")

statewide_trend <-
  covid_clean |>
  group_by(last_update_date) |>
  summarise(new_cases = mean(new_cases, na.rm = TRUE)) |>
  mutate(series = "Statewide Average")

combined_trend <- bind_rows(hartford_trend, statewide_trend)
```

::: {.callout-note}
`new_cases` is the day-over-day change in `total_cases`, computed with
`lag()` back in setup, grouped by town so one town's data never bleeds
into another's. See chapter 1's project for why this number can come
out negative on a day the state revises an earlier total — real
pandemic reporting data, this dataset included, is rarely clean enough
to avoid that.
:::

```{r}
#| label: trend-plot
ggplot(
  data = combined_trend,
  mapping = aes(x = last_update_date, y = new_cases, color = series)
) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Hartford" = "#3371E7",
      "Statewide Average" = "gray50"
    )
  ) +
  labs(
    title = "Hartford New Cases vs. Statewide Average",
    subtitle = "Daily count from report to report",
    x = NULL,
    y = "New Cases"
  ) +
  theme_classic()
```

Don't leave that read to the eye — state it. Average each series over
its last week of reports and compare:

```{r}
#| label: trend-finding
recent_avgs <-
  combined_trend |>
  filter(!is.na(new_cases)) |>
  group_by(series) |>
  slice_max(last_update_date, n = 7) |>
  summarise(avg_new = mean(new_cases), .groups = "drop")

hartford_avg <- recent_avgs$avg_new[recent_avgs$series == "Hartford"]
state_avg <- recent_avgs$avg_new[recent_avgs$series == "Statewide Average"]
standing <- if (hartford_avg > state_avg) "above" else "at or below"

recent_avgs
```

The line for the brief: over the last week of reports Hartford averaged
about `r round(hartford_avg, 1)` new cases a day against the statewide
`r round(state_avg, 1)` — running **`r standing`** the rest of the state.
The chart shows it; this sentence says it.

## Step 2 — Save the trend chart

```{r}
#| label: save-trend
#| eval: false
ggsave(
  here("outputs", "figures", "hartford_vs_state_trend.png"),
  width = 7,
  height = 4,
  dpi = 300
)
```

## Step 3 — This week's statewide map

The lesson's choropleth mapped `case_rate` — cumulative cases per
100,000 people. The Commissioner's packet already has that map from
chapter 3. This time, map a different question: **of everyone tested,
what share came back positive?** That's test positivity —
`number_of_positives / number_of_tests`, as a percentage. A town can
have a low case rate but a high positivity rate if it isn't testing
much; the two numbers tell different stories, and a briefing packet
usually wants both.

```{r}
#| label: map-data
ct_towns <-
  county_subdivisions(
    state = "09",
    cb = TRUE,
    progress_bar = FALSE
  )

latest_date <- max(covid_clean$last_update_date)

town_positivity <-
  covid_clean |>
  filter(last_update_date == latest_date) |>
  mutate(positivity = number_of_positives / number_of_tests * 100) |>
  select(town, positivity)

ct_bounded <-
  ct_towns |>
  left_join(town_positivity, by = join_by(NAME == town))
```

```{r}
#| label: map-plot
ggplot() +
  geom_sf(
    data = ct_bounded[is.na(ct_bounded$positivity), ],
    fill = "gray95",
    color = "white",
    linewidth = 0.1
  ) +
  geom_sf(
    data = ct_bounded[!is.na(ct_bounded$positivity), ],
    mapping = aes(fill = positivity),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_gradient2(
    low = "#C6D4FB",
    mid = "#7094F5",
    high = "#00214D",
    midpoint = mean(ct_bounded$positivity, na.rm = TRUE),
    na.value = "gray95",
    name = "Test Positivity\n(%)"
  ) +
  labs(
    title = "COVID-19 Test Positivity by Town",
    subtitle = paste("As of", latest_date),
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    legend.position = "right"
  )
```

## Step 4 — Save the map

```{r}
#| label: save-map
#| eval: false
ggsave(
  here("outputs", "figures", "statewide_positivity_map.png"),
  width = 7,
  height = 5,
  dpi = 300
)
```

## Step 5 — A quick top 5 / bottom 5 look

Not everything needs a map. Sometimes the Commissioner just wants a
short list: which five towns are worst, and which five are best.
`slice_max()` and `slice_min()` grab the top and bottom rows by a
column, no `arrange()` + `head()` needed.

```{r}
#| label: positivity-extremes
town_positivity |>
  slice_max(positivity, n = 5)

town_positivity |>
  slice_min(positivity, n = 5)
```

Two plain tibbles, printed as-is. Dressing a table like this up —
borders, color-coded cells, a footnote citing the data source — is
exactly what table-formatting packages like `gt` are built for. That's
further down the road; for now, a sorted tibble already answers the
question.

## Wrap-up

Two figures and a ranked list, all built from the same `covid_clean`
table chapter 2 handed you. That's the pattern this book is building
toward: clean once, reuse everywhere, and let each chapter's new skill
turn that clean data into something a person outside R can actually
read.

::: {.callout-note}
Want to compare against a reference? See
[`project-snapshots/after-ch-03`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/after-ch-03)
on GitHub for the `scripts/` this chapter should leave you with.
:::
