---
title: "Building a Dataset"
packages: [dplyr, tidyr, readr, janitor, stringr, lubridate]
---

```{r}
#| label: chapter-root
#| include: false
source(here::here("preamble-packages.R"))
source(here::here("preamble-data.R"))
```

```{r}
#| label: packages-header
#| echo: false
#| results: asis
packages_used_callout(rmarkdown::metadata$packages)
```

```{r}
#| label: load-linelist
#| include: false
linelist <- read_book_data("synthetic-covid-linelist.csv")
library(dplyr)
```

Every dataset so far has been a **count** table: one row per town per report
date, with cumulative cases and deaths. That's what the state publishes, and
it's plenty for trends and maps. But a lot of everyday epi work happens one
level down — at the **case** level. How old were the people who got sick? How
long between symptom onset and a positive report? What share of hospitalized
patients died, and did that differ by age?

None of those questions can be answered from a count table, because the
individual people have already been summed away. To practice case-level work
you need a **line-list**: one row per case, with that person's age, dates, and
outcomes.

Real line-lists exist — but they're confidential. You can't paste real
patients into a public book. So this chapter does the next best thing: it
**manufactures** a fake line-list from the real counts, and the act of
building it *is* the lesson. It also means stepping away from the town-level
Commissioner briefing for a chapter or two — the real town data resumes in the
Data Visualization chapter — while we drop down to the case level, where a
safe-to-share synthetic dataset is the only honest option. Along the way you'll
expand a count table into
individual rows, invent plausible values with controlled randomness, recode
and categorize variables, and do arithmetic on dates — the core moves of data
manipulation.

::: {.callout-warning}
## This data is fake

`synthetic-covid-linelist.csv` is **synthetic**. The town names and the number
of cases per town are real (from the CT Open Data Portal), but every age, sex,
race/ethnicity, date, hospitalization, and death is randomly generated. **No
row is a real person.** It exists only to practice on — never analyze it as if
it were real, and never mix it into real reporting.
:::

## The plan

We'll scope the work to **Hartford County during the winter 2020–2021 surge**
(December 2020 through January 2021) — a real wave, before vaccines were widely
available, and a chunk of data small enough to build on a laptop. The recipe
has three moves:

1. Start from the real town counts and work out how many *new* cases each town
   had on each report date.
2. Expand those counts into one row per case.
3. Attach invented-but-plausible demographics and outcomes to each row.

## Part A — How the line-list was built

The full recipe lives in `data-raw/make-synthetic-linelist.R` in the book's
repository. We'll walk through it a piece at a time. You don't need to run it
yourself — the finished file is in the [Data
Library](../data-library/index.qmd) — but reading it teaches the moves.

### Real counts into new cases

`Total cases` is cumulative, so a town's *new* cases on a given day are the
change from its previous report — the same `lag()` trick from earlier
chapters. We floor negatives at zero (the state sometimes revises an old total
downward, and you can't have negative people), then keep only our county and
our two-month window.

```{r}
#| label: recipe-counts
#| eval: false
hartford_county <- c(
  "Avon",
  "Berlin",
  "Bloomfield",
  "Bristol",
  "Burlington",
  "Canton",
  "East Granby",
  "East Hartford",
  "East Windsor",
  "Enfield",
  "Farmington",
  "Glastonbury",
  "Granby",
  "Hartford",
  "Hartland",
  "Manchester",
  "Marlborough",
  "New Britain",
  "Newington",
  "Plainville",
  "Rocky Hill",
  "Simsbury",
  "Southington",
  "South Windsor",
  "Suffield",
  "West Hartford",
  "Wethersfield",
  "Windsor",
  "Windsor Locks"
)

counts <-
  read_csv(
    here("data", "raw", "ct_covid_by_town_2026-06-30.csv")
  ) |>
  clean_names() |>
  mutate(
    report_date = mdy(last_update_date),
    town = str_to_title(str_trim(town)),
    total_cases = parse_number(as.character(total_cases))
  ) |>
  filter(town %in% hartford_county) |>
  arrange(town, report_date) |>
  mutate(new_cases = total_cases - lag(total_cases), .by = town) |> # .by is great so you don't forget to ungroup after doing a grouped mutate
  mutate(new_cases = pmax(new_cases, 0)) |>
  filter(
    report_date >= ymd("2020-12-01"),
    report_date <= ymd("2021-01-31"),
    !is.na(new_cases),
    new_cases > 0
  ) |>
  select(town, report_date, new_cases)
```

That leaves a compact table: for each town and report date, how many new cases
showed up. Now we blow it apart.

### One row per case with `uncount()`

`tidyr::uncount()` is the workhorse of this chapter. Hand it a column of
counts and it repeats each row that many times — the exact inverse of
`count()`. A row that says "Hartford, 2020-12-15, 90 new cases" becomes 90
identical Hartford rows.

```{r}
#| label: recipe-uncount
#| eval: false
linelist <-
  counts |>
  uncount(new_cases) |>
  mutate(case_id = sprintf("HC-%06d", row_number()))
```

Each row is now one case. They're still identical within a town-day, so the
last step is to give each one an identity.

### Inventing values with `set.seed()` and random draws

To fill in age, sex, and the rest we *draw* values at random from
distributions we pick. `sample()` draws from a set of categories with given
probabilities; `rnorm()` draws numbers from a bell curve; `rbinom()` flips
weighted coins (hospitalized or not, died or not).

Random draws are different every time you run them — which would make the
dataset impossible to reproduce. `set.seed()` fixes the random number
generator so every draw comes out the same on every run. Set it once, at the
top, before any drawing happens.

```{r}
#| label: recipe-seed
#| eval: false
set.seed(20210101)
```

::: {.callout-note}
## A cleaner seed is coming later

`set.seed()` changes the random state for the *rest of your session*, which is
fine in a one-shot script like this but can surprise you in a longer analysis.
The reproducibility chapter will introduce
[`withr::with_seed()`](https://withr.r-lib.org/), which sets a seed for just
one expression and then puts the random state back the way it found it — the
tidier tool when you only want one block of code to be reproducible. For a
standalone data-generation script, plain `set.seed()` at the top is perfectly
good.
:::

With the seed fixed, every field is one `mutate()` away. A few patterns worth
naming:

- **A number from a distribution** — age from a bell curve centered on 42,
  clamped to a human range with `pmin()`/`pmax()`:

  ```r
  age = pmin(pmax(round(rnorm(n, mean = 42, sd = 22)), 0L), 100L)
  ```

- **A category with set odds** — `sample()` with a `prob` vector. These
  proportions are illustrative, not measured:

  ```r
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.51, 0.49))
  ```

- **A weighted coin flip that depends on another column** — `case_when()`
  builds an age-dependent probability, then `rbinom()` flips it. Hospitalization
  risk climbs steeply with age:

  ```r
  hosp_prob = case_when(
    age < 18 ~ 0.01,
    age < 50 ~ 0.02,
    age < 65 ~ 0.06,
    age < 80 ~ 0.15,
    TRUE     ~ 0.30
  )
  hospitalized = if_else(rbinom(n, 1, hosp_prob) == 1, "Yes", "No")
  ```

- **A date relative to another date** — onset happens a few days *before* the
  report; subtract a random number of days:

  ```r
  onset_date = report_date - sample(3:14, n, replace = TRUE)
  ```

The full script layers a handful of these together, drops the intermediate
probability columns, and writes the result with `write_csv()`. That's the
whole trick: real counts decide *how many* rows, random draws decide *what's in
them*.

## Part B — Working with the line-list

Here's the finished product. In your own project you'd read the file you built:

```{r}
#| label: read-linelist-shown
#| eval: false
linelist <- read_csv(here("data", "synthetic-covid-linelist.csv"))
```

```{r}
#| label: glimpse-linelist
glimpse(linelist)
```

The twelve columns:

| Column | What it is |
|---|---|
| `case_id` | Unique fake ID, one per case |
| `town` | Hartford County town (real) |
| `age` | Age in years (simulated) |
| `sex` | `Female` / `Male` (simulated) |
| `race_ethnicity` | Race/ethnicity category (simulated) |
| `onset_date` | Symptom onset date (simulated, before the report) |
| `report_date` | Date the case was reported (real report date) |
| `hospitalized` | `Yes` / `No` (simulated) |
| `hosp_admit_date` | Admission date, or `NA` if not hospitalized |
| `hosp_discharge_date` | Discharge date, or `NA` |
| `died` | `Yes` / `No` (simulated) |
| `death_date` | Date of death, or `NA` |

Everything below runs on this real (fake) data.

### Categorizing a continuous variable

Age as a raw number is hard to tabulate — you rarely want a row for every
single year. Collapsing it into groups is a job for `case_when()`, reading top
to bottom and stopping at the first match:

```{r}
#| label: age-groups
linelist <- linelist |>
  mutate(
    age_group = case_when(
      age < 18 ~ "0-17",
      age < 50 ~ "18-49",
      age < 65 ~ "50-64",
      age < 80 ~ "65-79",
      TRUE ~ "80+"
    )
  )

count(linelist, age_group)
```

### Recoding values

`hospitalized` and `died` arrived as `"Yes"`/`"No"` text. For counting that's
fine, but for modelling later you'll often want a clean logical. `if_else()`
recodes a single variable in one line:

```{r}
#| label: recode-flags
linelist <- linelist |>
  mutate(
    hospitalized_flag = if_else(hospitalized == "Yes", TRUE, FALSE),
    died_flag = if_else(died == "Yes", TRUE, FALSE)
  )

count(linelist, hospitalized, hospitalized_flag)
```

### Doing arithmetic on dates

Subtracting one date from another gives a number of days. Two classic epi
measures fall straight out of that:

- **Reporting delay** — how long from symptom onset to the case being
  reported.
- **Length of stay** — how long a hospitalized patient was admitted.

```{r}
#| label: date-math
linelist <- linelist |>
  mutate(
    reporting_delay = as.integer(report_date - onset_date),
    length_of_stay = as.integer(hosp_discharge_date - hosp_admit_date)
  )

linelist |>
  summarise(
    median_delay = median(reporting_delay),
    median_stay = median(length_of_stay, na.rm = TRUE)
  )
```

`length_of_stay` is `NA` for anyone who was never admitted, which is exactly
right — there's no stay to measure. That's why `na.rm = TRUE` is needed in the
summary.

### Selecting a date range

Filtering to a window is the same `filter()` you already know, with date
comparisons. Say you only want cases reported in the first two weeks of
December:

```{r}
#| label: date-range
linelist |>
  filter(
    report_date >= as.Date("2020-12-01"),
    report_date <= as.Date("2020-12-14")
  ) |>
  nrow()
```

### A quick sanity check

Before trusting any built dataset, look for values that can't be real. Ages
should sit between 0 and about 120; onset should never come *after* the
report. A `filter()` that returns zero rows is a clean bill of health:

```{r}
#| label: range-check
linelist |>
  filter(age < 0 | age > 120 | onset_date > report_date) |>
  nrow()
```

Zero out-of-range rows — no surprise, since we built the data to behave, but on
real data this is the check that catches a birth year of 1850 or a typo'd date.

### Does it match the real counts?

The whole point of anchoring to real data is that the totals should line up.
The synthetic line-list has one row per real case in the window, so its row
count equals the real number of new cases Hartford County recorded from
December 2020 through January 2021:

```{r}
#| label: validate
nrow(linelist)
```

```{r}
#| label: validate-town
linelist |>
  count(town, name = "cases") |>
  arrange(desc(cases))
```

The demographics are invented, but the *shape* — which towns had the most
cases, when the wave peaked — is real, because the counts came from real data.

## Wrap-up

You started with a published count table and ended with a case-level
line-list you can slice by age, date, and outcome. More importantly you
practiced the manipulation moves that show up everywhere: `uncount()` to go
from counts to rows, `case_when()` to recode and categorize, seeded random
draws to fill in values reproducibly, and date arithmetic to turn two dates
into a meaningful number. The next chapters lean on exactly these skills — now
on data that finally has a person in every row.

::: {.callout-note}
The finished file, its source page, and the generator script are cataloged in
the [Data Library](../data-library/index.qmd).
:::

::: {.callout-note}
Want the finished generator in one file? See
[`project-snapshots/line-list-practice`](https://github.com/CT-DPH-Data-Management-and-Governance/modern-r-for-public-health-practitioners/tree/main/project-snapshots/line-list-practice)
on GitHub. `01_make_synthetic_linelist.R` is the finished generator this chapter
walks through — drop it into your project's `scripts/` (or `data-raw/`) folder
and run it to build the line-list yourself.
:::
